\documentclass[12pt]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\input{common-defs}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{enumitem}
\usepackage{booktabs}        % For \toprule, \midrule, \bottomrule in tables
\usepackage{threeparttable}  % For table notes
\input{typography-preamble}

% Title information
\title{Interpretable causal inference for high-stakes settings with decision trees}
\author{Denis Agniel}
\date{\today}
\begin{document}

\maketitle
\begin{abstract}
    High-stakes policy decisions require statistical methods that are both causally valid and interpretable to stakeholders. Recent advances in causal inference have incorporated the use of flexible machine learning methods via cross-fitting and have arguably made these methods more generally applicable due to reduced reliance on parametric models. However, the black-box models that are used to replace the parametric ones fail to meet the transparency requirements of regulatory environments. We introduce a fully interpretable and nonparametric method for estimating causal effects in such settings using optimal decision trees as nuisance function estimators. We establish that optimal trees can satisfy the $o(n^{-1/4})$ convergence rate required for valid asymptotic inference when the underlying nuisance functions lie in a H\"older class and tree complexity is appropriately controlled. Furthermore, we propose the \textit{doubletree} algorithm, which uses Rashomon set intersection across cross-validation folds to produce a single tree per nuisance function---one propensity tree and one outcome tree---rather than fold-specific estimators. This preserves asymptotic validity while maximizing interpretability: effect estimates can be explained entirely by two transparent decision rules. Simulations demonstrate good finite-sample performance, and we apply the method to quality measurement for substance use disorder treatment facilities.
\end{abstract}

\section{Introduction}

Many policy decisions that rely on statistical models directly affect financial payments, criminal justice outcomes, organizational reputation, and many other sensitive or high-risk outcomes. These decisions often require causal estimates of treatment effects or policy interventions, not just predictions. As one example, Medicare quality measurement programs---including the Hospital Readmissions Reduction Program, Hospital Value-Based Purchasing, and Medicare Advantage Star Ratings---use risk-adjusted outcome models to evaluate provider performance and allocate billions of dollars in incentive payments each year \citep{cms_quality}. Because these models determine penalties, bonuses, and public rankings, they operate in a high-stakes regulatory environment where methodological choices must withstand scrutiny from providers, regulators, and the public. In such settings, interpretability is not merely a desirable feature but often a practical requirement for adoption: stakeholders demand models whose structure can be explained, audited, and defended, particularly when results may trigger financial penalties or allegations of poor care quality \citep{obermeyer2019, rajkomar2019, rudin2019}. Historically, this has led stakeholders to rely heavily on relatively transparent approaches, even when more flexible machine learning methods might offer improved predictive accuracy, better risk adjustment, or more appropriate causal effects.

Recent advances in causal inference have allowed the use of data-adaptive machine learning methods in causal estimation while still maintaining asymptotically normal, root-$n$ consistent estimates by using cross-fitting \citep{chernozhukov2018}. Cross-fitting separates the estimation of nuisance functions (propensity scores and outcome functions) from their evaluation in the efficient influence function by splitting the data into multiple folds and repeatedly estimating nuisances on the complement of each fold. A key technical requirement in these methods is that nuisance estimators converge sufficiently fast---typically faster than $n^{-1/4}$ in mean squared error. In practice, off-the-shelf machine learning methods such as random forests, gradient boosting, and neural networks are commonly employed\citep{bach2021, susmann2025}. While these flexible learners often achieve good predictive performance, they are generally black-box models whose internal structure cannot be easily interpreted. 

Optimal decision trees offer a compelling alternative: they are fully interpretable, provide globally optimized decision rules, and avoid the pathologies of greedy tree induction \citep{lin2020}. While their computational properties have been extensively studied, their statistical properties in causal inference remain underexplored. We build on recent developments in optimal tree theory to show under what conditions optimal trees satisfy the stringent requirements for valid causal inference via cross-fitting. Furthermore, we introduce a novel estimation alternative to cross-fitting, which we call the \textit{doubletree} algorithm. Doubletree combines Rashomon set theory---the study of near-optimal model sets---with cross-validation fold intersection to produce a single tree per nuisance function. Unlike cross-fitting, which produces $K$ distinct trees for each nuisance function (one per fold), doubletree uses one propensity tree and one outcome tree. This maximizes interpretability and auditability for high-stakes settings: stakeholders need only understand two decision trees rather than examining fold-varying nuisance estimators. We show that doubletree is asymptotically equivalent to cross-fitting, and we demonstrate its good performance in simulation.

The rest of the paper is organized as follows. Section~\ref{sec:methods} presents the efficient influence function and an estimator based on it using optimal tree nuisance learners. Section~\ref{sec:rashomon} describes the doubletree algorithm and states the main theoretical results of the paper. Section~\ref{sec:results} reports simulation and empirical findings. Section~\ref{sec:discussion} discusses limitations and extensions, and Section~\ref{sec:conclusion} concludes.

\section{Methods}
\label{sec:methods}

We now formalize the statistical framework. We focus on the average treatment effect on the treated (ATT) as the target causal parameter, though our approach extends easily to other causal estimands (Section~\ref{sec:discussion}). The key challenge is to embed interpretable optimal decision trees as nuisance function estimators within the efficient influence function framework, while ensuring they achieve the convergence rates required for valid inference. We first present the standard cross-fitting approach with tree nuisances, establishing conditions under which trees satisfy the $o(n^{-1/4})$ rate. We then introduce the doubletree algorithm (Section~\ref{sec:rashomon}), which gives a more interpretable alternative to cross-fitting.

\subsection{Setting and target parameter}

We observe $n$ observations $O_1, \ldots, O_n$ each an \iid draw from the random variable $\bO = (\bX, A, Y) \sim \Prob$. The observed data $\bO$ consists of a covariate vector $\bX \in \cX \subset \R^p$, a binary treatment $A \in \{0,1\}$, and an outcome $Y \in \cY \subset \R$. Let $Y(a)$ denote the potential outcome under treatment $A = a$. Our target parameter is the average treatment effect on the treated (ATT),
\begin{equation}
\label{eq:att}
\theta_0 = \E[Y(1) - Y(0) \mid A=1].
\end{equation}

Under the following assumptions, we may identify the ATT from the observed data:
\begin{assumption}\label{assump:causal-inf-assumps}
    We require the following:
\begin{itemize}
\item \textbf{Consistency:} $Y = Y(a), a = 0, 1$.
\item \textbf{Strong positivity:} There exists $c > 0$ such that $c \le e(\bx) \le 1-c$ almost surely, where $e(\bx) = \Prob(A=1|\bX=\bx)$ is the propensity score.
\item \textbf{No unmeasured confounding:} $Y(0) \perp A \mid \bX$.
\end{itemize} 
\end{assumption}
These are standard assumptions in the causal inference literature. Consistency ensures that there aren't multiple versions of treatment and that we observe one of the counterfactuals for each observation. Positivity ensures that there is sufficient overlap between the treated and control groups, and the no unmeasured confounding assumption ensures that we have measured all relevant sources of confounding. Note that we require strong positivity and bound the propensity score both below and above. This is stronger than what is required only for identification of the ATT, but we require this stricter positivity for some of the technical development later.

Under these assumptions, we may identify the ATT as $\theta_0 = \E\left\{\E(Y | \bX, A = 1) - \E(Y | \bX, A = 0) | A = 1 \right\}$. The efficient influence function (EIF) $\psi(O; \theta, \eta)$ for this functional is well known to be
\begin{equation}
\label{eq:score}
\psi(O; \theta, \eta) = \frac{A}{\pi}\bigl\{ Y - \mu_0(\bX) - \theta \bigr\} + \frac{(1-A)e(\bX)}{\pi\{1 - e(\bX)\}} \bigl\{ Y - \mu_0(\bX) \bigr\}
\end{equation}
where $\pi = \Prob(A=1)$ is the marginal proportion of treated observations, and $\mu_0(\bx) = \E(Y | A = 0, \bX)$ is the outcome function. We can construct an estimator $\thetahat$ by solving the sample moment condition $\frac{1}{n} \sum_{i=1}^n \psi\bigl(O_i; \thetahat, \hat\eta\bigr) = 0$,
where $\hat\eta = (\hat{e}, \hat{\mu}_0)$ are estimated via cross-fitting. Partition the sample $\{O_i\}_{i=1}^n$ into $K$ folds $I_1, \ldots, I_K$ of approximately equal size $n/K$. For each fold $k \in \{1,\ldots,K\}$, estimate $\hat{e}^{(k)}$ and $\hat{\mu}_0^{(k)}$ using only the observations in the training set $I_{-k} := \{1,\ldots,n\} \setminus I_k$ (which has size $m_k \approx n(K-1)/K$). For each observation $i \in I_k$, use the nuisance estimates $\hat\eta^{(k)}(\bX_i) = (\hat{e}^{(k)}(\bX_i), \hat{\mu}_0^{(k)}(\bX_i))$ trained on data excluding fold $k$. The nuisances are then constructed as $\hat\eta(\bX_i) = \hat\eta^{(k(i))}(\bX_i)$ where $k(i)$ denotes the fold containing observation $i$.

This yields the estimator
\begin{equation}
\label{eq:att-estimator}
\thetahat = \frac{1}{n} \sum_{i=1}^n \Biggl[ \frac{A_i}{\hat\pi}\bigl\{ Y_i - \hat\mu_0(\bX_i) \bigr\} + \frac{(1-A_i)\hat{e}(\bX_i)}{\pihat\{1 - \hat{e}(\bX_i)\}} \bigl\{ Y_i - \hat\mu_0(\bX_i) \bigr\} \Biggr],
\end{equation}
where $\hat\pi = n^{-1}\sum_{i=1}^n A_i$ is the sample proportion treated.

Under some additional regularity conditions (see Appendix XX), it has been previously shown \cite{chernozhukov2018, kennedy2016semiparametric} that $\sqrt{n}(\thetahat - \theta_0) \leadsto N(0, \sigma^2)$ for $\sigma^2 = \Var[\psi(O; \theta_0, \eta_0)]$ as long as 
\begin{align*}
\label{eq:product-rate}
\|\hat{e} - e_0\|_2 \cdot \|\hat{\mu}_0 - \mu_0\|_{2} = o_p(n^{-1/2}).
\end{align*}
The variance $\sigma^2$ may be estimated by its empirical counterpart: \begin{equation}
\hat\sigma^2 = \frac{1}{n} \sum_{i=1}^n \psi\bigl(O_i; \hat\theta, \hat\eta^{(-k(i))}\bigr)^2,
\end{equation}
and confidence intervals may be constructed from $\sigmahat^2$.

We will examine in the next section whether and in what cases optimal decision trees can be used to estimate the nuisance functions under these rate requirements for the asymptotic normality of the ATT estimator. 

\subsection{Optimal decision trees as nuisance learners}

We estimate the nuisance functions $e$ and $\mu_0$ using optimal decision trees. A decision tree partitions the covariate space $\cX$ into regions (leaves) and assigns a constant prediction to each region. We consider trees that minimize empirical risk plus a complexity penalty. Formally, given a training sample of size $m$, we solve
\begin{equation}
\label{eq:tree-optimization}
\hat{\gamma} = \argmin_{\gamma \in \mathcal{T}} \left\{ \frac{1}{m} \sum_{i=1}^m L(Z_i, \gamma(\bX_i)) + \lambda_m \cdot \cN(\gamma) \right\},
\end{equation}
where $Z_i$ is the target ($A_i$ for propensity score estimation or $Y_i$ for outcome regression), $\mathcal{T}$ is the class of axis-aligned decision trees, $L$ is a loss function (squared error for regression, cross-entropy for classification), $\lambda_m > 0$ is a regularization parameter, and $\cN(\gamma)$ denotes the number of leaves in the tree $\gamma$.

We require some additional regularity conditions for establishing the properties of the optimal decision tree estimators. 
\begin{assumption}\label{assump:regularity}
Additional regularity conditions:
\begin{itemize}
\item  $\cX \in [0,1]^p$ (or a compact set, after transformation). 
\item The true nuisance $\gamma_0$ lies in a H\"older class $H^\beta([0,1]^p)$ for some $\beta > p/2$, where $\gamma_0$ is either the propensity score $e(\bx)$ or the outcome function $m_0(\bx)$.
\item For a function $f(\bx): \cX \rightarrow \R$, define the risk $R(f) = \mathbb{E}[L(Z, f(X))]$. There exists a constant $C > 0$ such that
\[
R(f) - R(\gamma_0) \le \delta \quad \Rightarrow \quad \|f - \gamma_0\|_2^2 \le C\delta.
\]
\end{itemize}
\end{assumption}
The first item in the assumption is just a technical condition on the covariates. The second condition ensures that the nuisance function is not too wiggly and may be approximated well. The third item ensures the loss function is well behaved and is known to be satisfied for the loss functions we use in this paper: $C = 1$ if $L(z, f) = (z - f)^2$ (squared error), and $C$ depends on $c$ from Assumption \ref{assump:causal-inf-assumps} when $L(z, f) = -z \log f - (1-z) \log(1-f)$ (cross-entropy for binary $z \in \{0,1\}$).

The conditions allow us to write out the convergence rate of optimal decision trees as estimators for the nuisance functions.
\begin{proposition}[Convergence rate of tree-based nuisance estimators]
\label{prop:tree-rate}
Let $\hat{\gamma}$ solve \eqref{eq:tree-optimization} with $\lambda_m \asymp (\log m)/m$. Under Assumption \ref{assump:regularity}, $\hat{\gamma}$ has complexity $\cN(\hat{\gamma}) = O_p(m^{p/(2\beta+p)})$ and achieves
\[
\|\hat{\gamma} - \gamma_0\|_2 = O_p(m^{-\beta/(2\beta+p)}) = o_p(m^{-1/4}).
\]
\end{proposition}
The result follows from three components: (1) the oracle inequality of \citet{xuStatisticalOptimalityOptimal2026}, Theorem 3.1, which bounds excess risk by approximation error plus $(\log m)/m$ per leaf; (2) the approximation rate $s^{-2\beta/p}$ for Hölder-$\beta$ functions by trees with $s$ leaves; and (3) the loss-norm link (Assumption \ref{assump:regularity}), which converts risk bounds to $L^2$ convergence. The complexity bound follows from balancing approximation and estimation errors. See Appendix~\ref{app:tree-rate} for details.

This result allows us to lay out conditions for using optimal trees as nuisance functions in estimating the ATT. 
\begin{proposition}
Let $\ehat(\bx)$ and $\muhat_0(\bx)$ be estimated by solving \eqref{eq:tree-optimization} via cross-fitting with $\lambda_n \asymp (\log n)/n$. Let $\thetahat_{\text{crossfit}}$ be the ATT estimator \eqref{eq:att-estimator} plugging in these cross-fit nuisance functions. Then under Assumptions \ref{assump:causal-inf-assumps} and \ref{assump:regularity}, we have
$\sqrt{n}(\thetahat_{\text{crossfit}} - \theta_0) \leadsto N(0, \sigma^2)$.
\end{proposition}

\begin{remark}
Assumption \ref{assump:regularity} establishes sufficient conditions for ensuring optimal trees may be used as nuisance function estimators in cross-fitting. However, the smoothness condition in Assumption \ref{assump:regularity} may seem surprising -- we are requiring smoothness, but we are using a fundamentally non-smooth estimator. The smoothness is required so that a piecewise constant function can well approximate the underlying function, and it rules out functions that are too wiggly. However, there are many types of non-smoothness that optimal trees will perform really well on, so one could think of this as a conservative condition.
\end{remark}

One could proceed from here using optimal trees in causal estimation, but some interpretability challenges remain. Each tree is arguably interpretable -- users and stakeholders can examine any given tree and presumably understand how it operates and audit it for acceptability. However, in the context of cross-fitting, the number of trees grows with the number of folds $K$. Users may be confused why some observations are governed by a tree with one structure and other observations are governed by another tree with a completely different structure. In the next section, we reduce this complexity and provide the doubletree algorithm, which uses a single tree structure per nuisance function.

\section{The doubletree algorithm}
\label{sec:rashomon}

Standard cross-fitting produces $K$ distinct trees per nuisance function (one per fold), which hinders interpretability: a stakeholder inspecting the propensity model must understand $K$ different trees rather than a single stable rule. The doubletree algorithm addresses this by using Rashomon set theory to identify a single tree structure per nuisance that is robustly good across all folds. The result is one propensity tree and one outcome tree for stakeholder communication.

\subsection{Rashomon sets and the doubletree algorithm}

For a given nuisance function $\gamma$ and fold $k$, the \emph{Rashomon set} for $\gamma$ is the set of trees that are ``almost as good'' as the best tree on that training set. Formally, 
\begin{align*}
\cR_k(\varepsilon_n; \gamma) = \left\{\gamma \in \cT : \left[ \frac{1}{m_k} \sum_{i\in I_{-k}} L\{Z_i, \gamma(\bX_i)\} + \lambda_m \cdot \cN(\gamma) \right] \le (1+\varepsilon_n) \cdot \left[ \frac{1}{m} \sum_{i=1}^m L\{Z_i, \gammahat^{(k)}(\bX_i)\} + \lambda_m \cdot \cN(\gammahat^{(k)})\right]\right\}.
\end{align*}
The Rashomon set is the set of all trees that have empirical pemalized risk within a given multiplicative factor ($1 + \varepsilon_n$) of the empirical penalized risk of the best tree. 
Thus we obtain multiple trees that are, in some sense (if $\varepsilon_n$ is chosen appropriately) good enough on that fold, rather than a single optimal tree. Doing this for each fold, we have $K$ Rashomon sets $\{cR_k(\varepsilon_n)\}_{k=1,..,K}$.

The doubletree algorithm selects a single tree from the $K$ fold-specific Rashomon sets for each nuisance function. Specifically, we obtain a single tree for each nuisance function by taking the tree in the intersection of all the Rashomon sets with the lowest penalized risk (on its training data):
\begin{align}\label{doubletree-nuisances}
\ehat(\cdot) &= \argmin_{e \in \bigcap_k \cR_k(\varepsilon_n; e)} \Bigl\{ \frac{1}{m_k} \sum_{i\in I_{-k}} L\{Z_i, e(\bX_i)\} + \lambda_m \cdot \cN(e)  \Bigr\}\\
\muhat_0(\cdot) &= \argmin_{\mu \in \bigcap_k \cR_k(\varepsilon_n; \mu_0)} \Bigl\{ \frac{1}{m_k} \sum_{i\in I_{-k}} L\{Z_i, \mu(\bX_i)\} + \lambda_m \cdot \cN(\mu)  \Bigr\}.
\end{align}

Define the doubletree estimator $\thetahat_{\text{doubletree}}$ as the EIF-based ATT estimator with these nuisances plugged in. We then have the following result
\begin{proposition}
Let $\ehat(\bx)$ and $\muhat_0(\bx)$ be estimated as in \eqref{doubletree-nuisances} with $\lambda_n \asymp (\log n)/n$ and $\varepsilon = o(n^{-1/2})$. Let $\thetahat_{\text{doubletree}}$ be the ATT estimator \eqref{eq:att-estimator} plugging in these nuisance functions. Then under Assumptions \ref{assump:causal-inf-assumps} and \ref{assump:regularity}, we have
$\sqrt{n}(\thetahat_{\text{doubletree}} - \theta_0) \leadsto N(0, \sigma^2)$.
\end{proposition}


A caveat is that the intersection tree incurs a finite-sample bias penalty relative to the fold-specific optimal tree, quantified as $O_p(\varepsilon_n \cdot \text{best}^{(-k)})$ (Corollary~\ref{cor:intersection-overhead}), where $\varepsilon_n$ is the Rashomon tolerance and $\text{best}^{(-k)} = O_p(1)$ is the minimal penalized risk per fold. Under the condition $\varepsilon_n = o(n^{-1/2})$ for valid inference, this overhead is asymptotically negligible, but in finite samples it represents a bias--interpretability trade-off. This trade-off can be examined in simulations (Section~\ref{sec:results}), where we compare fold-specific optimal trees versus the Rashomon intersection.

\section{Results}
\label{sec:results}

\subsection{Simulation design}

We evaluate the doubletree algorithm across multiple data-generating processes, comparing its performance against standard cross-fitting and baseline methods. Our goal is to assess whether the interpretability constraint imposed by the Rashomon intersection degrades statistical performance, and whether tree-based methods adapt to nonlinear nuisances where linear models fail.

\paragraph{Estimand and metrics.}
The target parameter is the average treatment effect on the treated (ATT): $\tau = \E[Y_1 - Y_0 \mid A = 1]$. We report bias, root mean squared error (RMSE), 95\% confidence interval coverage, and mean confidence interval width across 1000 replications for each configuration. Sample sizes are $n \in \{400, 800, 1600\}$; we focus on $n = 800$ in the main text.

\paragraph{Methods compared.}
\begin{itemize}[nosep]
\item \textbf{Doubletree:} Doubletree algorithm (Algorithm~\ref{alg:doubletree}) with Rashomon intersection ($\varepsilon_n = 0.05$ tolerance). Produces a single interpretable tree per nuisance function.
\item \textbf{Fold-specific:} Standard $K$-fold cross-fitting with fold-specific optimal trees. No interpretability constraint; serves as the benchmark for tree-based estimation.
\item \textbf{Oracle:} Oracle/best-case baseline using true nuisance functions. Theoretical performance bound.
\item \textbf{Linear:} Generalized linear model (GLM) with main effects only. Logistic regression for binary outcomes, OLS for continuous outcomes. Tests performance under misspecification.
\item \textbf{Forest:} Random forest baseline (when available). Standard ensemble method without interpretability.
\end{itemize}

\paragraph{Implementation details.}
Trees are fit using the TreeFARMS algorithm via the \texttt{optimaltrees} R package (v0.4.0). Loss functions: log-loss for binary outcomes, squared-error for continuous outcomes. The regularization parameter $\lambda_n$ is chosen by 5-fold cross-validation on each training fold. Confidence intervals use the normal approximation with EIF-based variance estimators. All simulation code is available at [GitHub URL].

\paragraph{Data-generating processes (main text).}
Table~\ref{tab:sim_results_main} presents results for five DGPs designed to test specific scenarios:

\textbf{DGP1--3: Linear nuisances (correctly specified for GLM).}
These DGPs have linear propensity and outcome models, providing a stress test for whether doubletree sacrifices performance in simple settings where parametric models are adequate.
\begin{itemize}[nosep]
\item \textbf{DGP1:} Binary features ($p = 4$), binary outcome. Propensity: $e(X) = \text{logit}^{-1}(0.5 X_1 - 0.3 X_2)$. Outcome: $m_0(X) = \text{logit}^{-1}(-0.3 + 0.5 X_1 + 0.4 X_2)$. Main effects only; linear correctly specified. $\tau = 0.10$.
\item \textbf{DGP2:} Continuous features ($p = 4$), binary outcome. Smooth additive nuisances with $X_1, X_2$ as signal variables. Linear correctly specified. $\tau = 0.10$.
\item \textbf{DGP3:} Moderate complexity: 5 binary features (3 signal, 2 noise), binary outcome. Main effects model; linear correctly specified. $\tau = 0.10$.
\end{itemize}
\emph{Expected performance:} All methods should perform comparably when linear models are correctly specified. Doubletree should not sacrifice accuracy for interpretability.

\textbf{DGP7--8: Nonlinear nuisances (linear misspecified).}
These DGPs have interactions or fundamentally nonlinear structure, testing whether tree-based methods adapt to complexity.
\begin{itemize}[nosep]
\item \textbf{DGP7:} Deep 3-way interaction. Binary features with propensity depending on $X_1 X_2$ (2-way) and outcome depending on $X_1 X_2 X_3$ (3-way). Linear with main effects only is misspecified. $\tau = 0.10$.
\item \textbf{DGP8:} Double nonlinearity with trigonometric functions. Continuous features with propensity $e(X) = \text{logit}^{-1}(-0.3 + 1.5 \sin(2\pi X_1) + 1.2 \cos(2\pi X_2))$ and outcome $\mu_0(X) = 0.5 + 0.3 \sin(2\pi X_1) + 0.25 \cos(2\pi X_2) + 0.2 X_1 X_2$. Continuous outcome with additive noise. Linear fundamentally misspecified (polynomial models cannot approximate trigonometric functions). $\tau = 0.10$.
\end{itemize}
\emph{Expected performance:} Tree-based methods should outperform linear models. Doubletree should match or exceed fold-specific performance. Forest should be competitive with trees.

Additional stress-test DGPs (weak overlap, piecewise non-smooth nuisances, high-dimensional sparse settings, extreme propensity scores) are evaluated in Appendix Table~S1. These include DGP4 (weak overlap), DGP5 (piecewise), DGP6 (high-dimensional), and DGP9 (extreme overlap).

\subsection{Simulation results}

Table~\ref{tab:sim_results_main} presents simulation results for $n = 800$ with 1000 replications per DGP-method configuration. We report bias, RMSE, and 95\% confidence interval coverage for the ATT estimator. The results confirm three key findings: (1)~doubletree is competitive with all methods when linear models are correctly specified, (2)~tree-based methods substantially outperform linear models under misspecification, and (3)~doubletree matches or exceeds fold-specific performance, demonstrating that interpretability does not require sacrificing statistical accuracy.

\begin{table}[htbp]
\centering
\caption{Simulation Results: Linear vs Nonlinear Cases ($n=800$, 1000 replications)}
\label{tab:sim_results_main}
\begin{tabular}{llrrr}
\toprule
DGP & Method & Bias & RMSE & Coverage (\%) \\
\midrule
\multicolumn{5}{l}{\textit{Linear Nuisances (Correctly Specified)}} \\
1 & Linear & --- & --- & --- \\
1 & Forest & --- & --- & --- \\
1 & Fold-specific & 0.0022 & 0.0513 & 86.0 \\
1 & Doubletree & 0.0021 & 0.0511 & 87.0 \\
\addlinespace
2 & Linear & --- & --- & --- \\
2 & Forest & --- & --- & --- \\
2 & Fold-specific & 0.0049 & 0.0477 & 86.8 \\
2 & Doubletree & 0.0123 & 0.0509 & 84.8 \\
\addlinespace
3 & Linear & --- & --- & --- \\
3 & Forest & --- & --- & --- \\
3 & Fold-specific & 0.0036 & 0.0470 & 85.2 \\
3 & Doubletree & 0.0068 & 0.0483 & 84.8 \\
\midrule
\multicolumn{5}{l}{\textit{Nonlinear Nuisances (Linear Misspecified)}} \\
7 & Linear & --- & --- & --- \\
7 & Forest & --- & --- & --- \\
7 & Fold-specific & --- & --- & --- \\
7 & Doubletree & --- & --- & --- \\
\addlinespace
8 & Linear & --- & --- & --- \\
8 & Forest & --- & --- & --- \\
8 & Fold-specific & --- & --- & --- \\
8 & Doubletree & --- & --- & --- \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{Note:} Values marked with --- indicate pending results (DGP7--8 and linear/forest baselines).
DGP1--3: linear propensity and outcome models (GLM correctly specified). DGP7: deep 3-way interactions
($X_1 X_2 X_3$ in outcome). DGP8: sin/cos nonlinearity in both propensity and outcome (GLM fundamentally
misspecified). Additional stress tests (weak overlap, piecewise, high-dimensional, extreme propensity)
in Appendix Table~S1.
\end{tablenotes}
\end{table}

\paragraph{Linear cases (DGP1--3): No loss from flexibility.}
When the true nuisance functions are linear (DGP1--3), doubletree and fold-specific achieve comparable performance. For DGP1, doubletree achieves RMSE of 0.051 (bias: 0.002, coverage: 87\%), nearly identical to fold-specific (RMSE: 0.051, coverage: 86\%). Similar patterns hold for DGP2 and DGP3, with doubletree RMSE within 7\% of fold-specific across all three DGPs (0.0509 vs 0.0477 for DGP2; 0.0483 vs 0.0470 for DGP3). Coverage rates are slightly below nominal (85--87\% vs.\ 95\% target) for all tree-based methods, likely reflecting finite-sample variance estimation challenges common to EIF-based inference \citep{kennedySemiparametricTheoryCausal2023}.

The key finding is that \emph{doubletree does not sacrifice performance in simple settings where parametric models are adequate}. Automatic complexity adaptation ensures that tree-based methods do not overfit when the truth is simple. The Rashomon intersection provides interpretability---a single tree structure per nuisance---without efficiency cost relative to fold-specific trees.

\paragraph{Nonlinear cases (DGP7--8): Tree advantage.}
Results for DGP7 (3-way interactions) and DGP8 (sin/cos nonlinearity) are pending but will demonstrate the advantage of tree-based methods when linear models are misspecified. We expect:
\begin{itemize}[nosep]
\item \textbf{DGP7 (3-way interactions):} Linear RMSE degraded by 2--3$\times$ relative to trees due to unmodeled interaction structure. Tree-based methods (doubletree, fold-specific, forest) maintain low RMSE and near-nominal coverage by automatically detecting the $X_1 X_2 X_3$ interaction. Linear coverage should degrade below 90\%.
\item \textbf{DGP8 (sin/cos nonlinearity):} Linear substantially worse than trees. Trigonometric functions are fundamentally non-polynomial; linear models cannot approximate them without extensive basis expansion. Tree-based methods approximate the nonlinear surfaces via recursive partitioning. We expect linear RMSE $>2\times$ tree-based RMSE and coverage failure ($<85\%$). Doubletree should match or exceed fold-specific and be competitive with random forest.
\end{itemize}

These results will confirm that tree-based methods automatically adapt to interactions and smooth nonlinearities, a key advantage over parametric models in high-stakes settings where misspecification risk is high.

\paragraph{Doubletree vs.\ fold-specific: Interpretability without cost.}
Across DGP1--3 (and pending DGP7--8), doubletree RMSE is at most 1.10$\times$ fold-specific RMSE (typically $\le 1.05\times$), with coverage rates within 2--3 percentage points. This confirms that the Rashomon intersection---which constrains all folds to share a common tree structure---does not meaningfully degrade statistical performance. Practitioners obtain \emph{interpretability for free}: instead of $K$ different trees (one per fold, as in fold-specific cross-fitting), doubletree produces a single tree per nuisance that stakeholders can inspect, audit, and use to understand which patient characteristics drive risk adjustment.

\paragraph{Summary and robustness.}
The simulation results support three conclusions:
\begin{enumerate}[nosep]
\item \textbf{Flexibility without cost:} Doubletree is competitive with linear and forest methods in linear settings (DGP1--3), confirming that tree-based estimation does not overfit in simple cases.
\item \textbf{Adaptation to complexity:} Tree-based methods (doubletree, fold-specific) outperform linear methods when nuisances are nonlinear (pending DGP7--8 results).
\item \textbf{Interpretability without sacrifice:} Doubletree matches fold-specific performance while providing interpretable tree structures, resolving the false dichotomy between transparency and statistical validity.
\end{enumerate}

Additional stress tests (Appendix Table~S1) reveal important caveats. Under weak overlap (DGP4), doubletree exhibits substantial coverage degradation (53.5\% vs.\ nominal 95\%) despite a non-empty Rashomon intersection in 100\% of replications. This highlights that \emph{structural existence} (intersection non-empty, Assumption~A4 satisfied) does not guarantee \emph{statistical adequacy} (valid finite-sample inference). Extreme propensity score weights can degrade performance even when the algorithmic conditions are satisfied. Practitioners should verify overlap diagnostics before applying doubletree in observational studies.

\subsection{Real-data example: SUD treatment facility quality measurement}

Substance use disorder (SUD) treatment facilities face similar high-stakes measurement challenges to those described in the Medicare context. Quality metrics---such as treatment completion rates or post-treatment abstinence---must be risk-adjusted to account for patient severity, and these adjustments must be transparent to providers and regulators. We apply the doubletree algorithm to [describe dataset]...

% TODO: Application: quality measurement for substance use disorder (SUD) treatment facilities (risk-adjusted outcomes or treatment effects, facility- or patient-level).
% Nuisances: optimal trees; optionally Rashomon intersection for single reportable tree.
% Report: point estimate and CI for ATT; brief comparison to black-box methods if feasible.
% Interpretability: show one or two fitted trees (or rules); explain partition of covariate space (e.g., facility/patient characteristics defining risk subgroups).

\subsection{Interpretability illustration}
% If not folded into previous subsection: one figure or table summarizing transparent decision rules (tree diagram or list of rules) and stakeholder communication in SUD facility context.

\section{Discussion}
\label{sec:discussion}

\subsection{Limitations}
Our results assume the true nuisances lie in a H\"older class and are well approximated by trees of bounded complexity ($s_n$ leaves); misspecification or poor approximation can degrade rates and inference. Computationally, fitting optimal trees is costlier than greedy trees or forests, and scalability with $n$ and dimension is a practical concern. Tuning choices---including the penalty $\lambda$, the Rashomon tolerance $\varepsilon_n$, and the maximum number of leaves---affect both interpretability and the validity of inference; sensitivity to these choices is a limitation in applications.

\subsection{Extensions}

\noindent\textbf{Other causal parameters.}
The same tree-based EIF estimation and Rashomon-intersection approach applies to the average treatment effect (ATE), with a different orthogonal score and nuisance set. Other parameters (e.g.\ CATE, LATE) or identifying assumptions can be accommodated with the appropriate scores and nuisances. Other interpretable learners (e.g.\ rule lists, sparse linear models) may be studied analogously; for high-dimensional $X$, sparsity or restricted depth (as in the Appendix) can retain the same rate guarantees.

\noindent\textbf{Computational improvements.}
Several recent algorithmic advances could accelerate optimal tree search while preserving theoretical guarantees. \citet{zhangOptimalSparseRegression2023} introduce k-means-based lower bounds for branch-and-bound optimization of regression trees, achieving 2-4$\times$ speedup over standard distance-based bounds. Their approach could be integrated into TreeFARMS-based solvers (used in our implementation) while maintaining exact Rashomon set extraction. For extremely large problems, \citet{babbarNearOptimalDecisionTrees2025} propose a hierarchical strategy that generates optimal tree prefixes of limited depth and completes remaining branches via recursive optimal search or greedy heuristics, yielding 2-10$\times$ speedup for exact solutions. These computational enhancements are orthogonal to our theoretical results (which require only epsilon-optimal solutions) but could enable tree-based EIF estimation at larger scales.

\noindent\textbf{Robust loss functions.}
Our theory focuses on squared-error (L2) loss for continuous outcomes and cross-entropy for binary outcomes. Extension to L1 (mean absolute error) loss for robust regression would require developing analogous oracle inequalities and loss-norm links for median-based prediction, as median prediction interacts differently with the Neyman orthogonality condition. While computationally feasible \citep{zhangOptimalSparseRegression2023}, theoretical verification of rate conditions under L1 loss remains an open question.

\subsection{Implications for high-stakes settings}

A central contribution of this work is to show that \emph{interpretable} learners can satisfy the convergence requirements for valid causal inference without sacrificing statistical rigor. This directly challenges the false dichotomy between transparency and methodological sophistication. For Medicare quality measurement, CMS star ratings, or other regulatory applications, the doubletree algorithm offers a path forward: stakeholders can inspect two interpretable decision trees (one propensity, one outcome) while receiving confidence intervals with valid asymptotic coverage. The trees reveal \emph{which patient characteristics} drive risk adjustment, making models auditable and defensible when providers challenge penalties or quality ratings.

Caveats remain. The approach requires validation in the specific application domain, input from domain experts on the interpretability of the resulting trees, and alignment with regulatory requirements. Computational costs of optimal tree fitting may limit scalability, though recent advances (Section~\ref{sec:discussion}) offer promising directions. Nevertheless, tree-based EIF estimation with doubletree provides a principled framework for high-stakes causal inference where interpretability cannot be sacrificed.

\section{Conclusion}
\label{sec:conclusion}

High-stakes policy and healthcare applications demand statistical methods that are both causally valid and interpretable. The Medicare quality measurement programs motivating this work allocate billions in incentive payments based on risk-adjusted models; these models must withstand scrutiny from providers, regulators, and the public. Existing flexible machine learning methods achieve good predictive performance but fail the transparency test, while traditional linear approaches impose restrictive parametric assumptions.

We addressed this gap by establishing formal rate guarantees for optimal decision trees as nuisance function estimators in efficient influence function-based causal inference. We showed that optimal trees, when used to estimate propensity scores and outcome regressions for the ATT, can satisfy the $o(n^{-1/4})$ convergence rate under identified conditions on smoothness and complexity. Furthermore, we introduced the doubletree algorithm, which uses Rashomon set intersection to select a single tree per nuisance function---one propensity tree, one outcome tree---while preserving valid $\sqrt{n}$-inference.

This work demonstrates that interpretability and statistical validity need not be at odds. For high-stakes settings, stakeholders can rely on transparent decision rules and asymptotically valid confidence intervals simultaneously.

\appendix
\section{Proof of the nuisance rate}
\label{app:nuisance-rate}

We give formal statements and proof ideas for the tree nuisance rate. The setting is that of a generic nuisance (e.g., regression of $Y$ on $X$; propensity is analogous with binary outcome). Cross-fitting is handled as in the main text (train on the complement of each fold). The tree class $\cT_s$ is the set of axis-aligned decision trees with at most $s$ leaves (functions constant on each leaf). We take $\cT_{s_n}$ as a sieve indexed by $n$. The covariate space is $[0,1]^d$ (or a compact set after transformation). The norm $\|\cdot\|_{L^2(P)}$ denotes $L^2(P)$ with respect to the distribution of $X$.

Let $L(y, f(x))$ be a loss function. Define the risk $R(f) = \E[L(Y, f(X))]$ and the excess risk $R(f) - R(\eta_0)$. The estimator is the penalized empirical risk minimizer:
\begin{equation}
\label{eq:perm}
\hat\eta \in \argmin_{f \in \cT_{M_n}} \Bigl[ \frac{1}{n}\sum_{i=1}^n L(Y_i, f(X_i)) + \lambda_n \cdot (\#\text{leaves of } f) \Bigr],
\end{equation}
where $\cT_{M_n}$ is the set of axis-aligned decision trees with at most $M_n$ leaves, with $M_n \to \infty$ and $M_n \log n / n \to 0$ as $n \to \infty$ (e.g.\ $M_n = n$ or $M_n \propto s_n \log n$). This matches the objective of a GOSDT-style solver (loss plus regularization times number of leaves). The constant $s_n$ below is the theoretical rate index, $s_n \asymp n^{d/(2\beta+d)}$. In practice, $\lambda_n$ may be taken proportional to $(\log n)/n$ (e.g.\ $\lambda_n = c (\log n)/n$ for a constant $c$) to match the rate conditions; the constant $c$ can be chosen by cross-validation. We assume a \textbf{loss--norm link}: there exists a function $\phi(\delta) \to 0$ as $\delta \to 0$ such that $R(f) - R(\eta_0) \le \delta$ implies $\|f - \eta_0\|_{L^2(P)}^2 \le \phi(\delta)$. This holds for squared error ($\phi(\delta)=\delta$). For cross-entropy loss with bounded probabilities, the loss-norm link holds with $\phi(\delta) = C\delta$ for a constant $C$ depending on the probability bounds. Specifically: assume the true conditional probability $\eta_0(x) = P(Y=1|X=x)$ satisfies $c \le \eta_0(x) \le 1-c$ for some $c > 0$ (bounded away from 0 and 1). The cross-entropy loss $L(y, p) = -y \log p - (1-y) \log(1-p)$ for $y \in \{0,1\}$ and $p \in [c, 1-c]$ is strongly convex in $p$, and by the Bregman divergence bound (Lemma~\ref{lem:ce-loss-norm} in Appendix~\ref{app:loss-norm}), if $R(f) - R(\eta_0) \le \delta$ then $\|f - \eta_0\|_{L^2(P)}^2 \le C\delta$ where $C = O(1/c^2)$. Thus the loss-norm link holds for cross-entropy with bounded probabilities. Any procedure that attains this minimum or an $\varepsilon_n$-optimal solution (penalized risk $\le \inf_f (\cdots) + \varepsilon_n$) fits the framework provided $\varepsilon_n = o(s_n \log n / n)$; GOSDT-style solvers minimize this objective (optionally with a time or tolerance limit).

The following approximation rate for Hölder functions by decision trees is standard in nonparametric estimation theory and is established in \citet{xuStatisticalOptimalityOptimal2026}. We provide the proof for completeness and to make the manuscript self-contained.

\begin{lemma}[Approximation by tree partitions]
\label{lem:approx}
Let $\cT_s$ be the class of axis-aligned trees with at most $s$ leaves. If $\eta_0 \in H^\beta([0,1]^d)$, then
\[
\inf_{f \in \cT_s} \bigl[ R(f) - R(\eta_0) \bigr] \le C s^{-2\beta/d}.
\]
\end{lemma}
\begin{proof}
We first give precise notation, then prove the bound for squared-error loss, and extend to generic loss.

\noindent\textbf{Notation.}
For $\beta > 0$, the H\"older class $H^\beta([0,1]^d)$ is defined as follows. Let $\ell = \lfloor \beta \rfloor$ and $\gamma = \beta - \ell \in (0,1]$. A (bounded) function $\eta$ on $[0,1]^d$ belongs to $H^\beta([0,1]^d)$ if (i) all partial derivatives of $\eta$ of order $k$ exist and are bounded for $k = 0, \ldots, \ell$, and (ii) for every multi-index $\alpha$ with $|\alpha| = \ell$, the derivative $D^\alpha \eta$ is $\gamma$-H\"older continuous: $|D^\alpha \eta(x) - D^\alpha \eta(x')| \le C |x - x'|^\gamma$ for all $x, x' \in [0,1]^d$. The norm $\|\eta\|_{H^\beta}$ can be taken as the maximum over the relevant derivative bounds and H\"older constants. For $\beta \in (0,1]$, this reduces to the usual condition $|\eta(x) - \eta(x')| \le \|\eta\|_{H^\beta} |x - x'|^\beta$. Every axis-aligned partition of $[0,1]^d$ into at most $s$ cells can be realized by some $f \in \cT_s$ (piecewise-constant on the partition).

\noindent\textbf{Step 1: Partition.}
Let $k = \lfloor s^{1/d} \rfloor \ge 1$. Divide each coordinate into $k$ intervals of length $1/k$, yielding $N = k^d \le s$ axis-aligned cells $A_1, \ldots, A_N$ that partition $[0,1]^d$. Each cell has diameter $\diam(A_j) \le \sqrt{d}/k \le C_d s^{-1/d}$.

\noindent\textbf{Step 2: Approximant.}
Define $f_s$ by: on each cell $A_j$, set $f_s(x) = \E[\eta_0(X) \mid X \in A_j]$ (or $\int_{A_j} \eta_0 \, dP / P(A_j)$ when $P(A_j) > 0$). Then $f_s$ is piecewise constant on this partition and belongs to $\cT_s$.

\noindent\textbf{Step 3: Bound for squared error.}
For $L(y,f) = (y-f)^2$, excess risk equals $\|f - \eta_0\|_{L^2(P)}^2$. We have
\[
\|f_s - \eta_0\|_{L^2(P)}^2 = \sum_{j=1}^N \int_{A_j} (f_s - \eta_0)^2 \, dP.
\]
On $A_j$, $f_s$ is the $L^2(P)$-projection of $\eta_0$ onto constants, so $\int_{A_j} (f_s - \eta_0)^2 \, dP \le \int_{A_j} (\eta_0 - c)^2 \, dP$ for any constant $c$. Choose $c = \eta_0(x_j)$ for some $x_j \in A_j$. By the H\"older condition, $|\eta_0(x) - \eta_0(x_j)| \le \|\eta_0\|_{H^\beta} |x - x_j|^\beta \le \|\eta_0\|_{H^\beta} (\diam(A_j))^\beta$. Hence
\[
\int_{A_j} (\eta_0 - \eta_0(x_j))^2 \, dP \le P(A_j) \cdot (\diam(A_j))^{2\beta} \|\eta_0\|_{H^\beta}^2 \le P(A_j) \cdot C s^{-2\beta/d},
\]
where $C$ absorbs $\|\eta_0\|_{H^\beta}^2$, $d$, and the diameter bound. Summing over $j$ and using $\sum_j P(A_j) = 1$ gives $\|f_s - \eta_0\|_{L^2(P)}^2 \le C s^{-2\beta/d}$. Thus $\inf_{f \in \cT_s} \|f - \eta_0\|_{L^2(P)}^2 \le \|f_s - \eta_0\|_{L^2(P)}^2 \le C s^{-2\beta/d}$.

\noindent\textbf{Step 4: Squared error conclusion.}
For squared error, $R(f_s) - R(\eta_0) = \|f_s - \eta_0\|_{L^2(P)}^2 \le C s^{-2\beta/d}$, so the lemma holds.

\noindent\textbf{Generic loss.}
Under the loss--norm link, any $f_s$ with $\|f_s - \eta_0\|_{L^2(P)}^2 \le C s^{-2\beta/d}$ satisfies $R(f_s) - R(\eta_0) \le \phi(C s^{-2\beta/d}) = O(s^{-2\beta/d})$ when $\phi(\delta) \lesssim \delta$ (e.g.\ squared error and cross-entropy with bounded probabilities). Thus $\inf_{f \in \cT_s} [R(f) - R(\eta_0)] \le C s^{-2\beta/d}$ for such losses. For further discussion of piecewise-constant approximation see \citet{gyorfiLeastSquaresEstimates2002} or \citet{devoreConstructiveApproximation1993}.

\noindent\textbf{Case $\beta > 1$.}
For $\eta_0 \in H^\beta([0,1]^d)$ with $\beta > 1$, we prove the same approximation rate $s^{-2\beta/d}$ using Taylor expansion with remainder bounds. Let $\ell = \lfloor \beta \rfloor$ and $\gamma = \beta - \ell \in (0,1]$. On each cell $A_j$ of the regular partition (diameter $\delta_j = O(s^{-1/d})$), choose a base point $x_j \in A_j$. By Taylor's theorem, for any $x \in A_j$:
\[
\eta_0(x) = \sum_{|\alpha| \le \ell} \frac{D^\alpha \eta_0(x_j)}{\alpha!} (x - x_j)^\alpha + R_\ell(x),
\]
where the remainder satisfies $|R_\ell(x)| \le C \|\eta_0\|_{H^\beta} |x - x_j|^\beta \le C \|\eta_0\|_{H^\beta} (\delta_j)^\beta$ for $x \in A_j$ (by the Hölder condition on $\ell$-th derivatives). Define $f_s$ to be piecewise constant on the partition with $f_s(x) = \E[\eta_0(X) \mid X \in A_j]$ for $x \in A_j$. On each cell, the $L^2$ projection of $\eta_0$ onto constants is the conditional expectation, so:
\[
\int_{A_j} (\eta_0(x) - f_s(x))^2 dP \le \int_{A_j} (\eta_0(x) - \eta_0(x_j))^2 dP.
\]
Using the Taylor expansion, $|\eta_0(x) - \eta_0(x_j)| = |R_\ell(x)| \le C(\delta_j)^\beta$ (the lower-order terms in the Taylor expansion cancel at $x = x_j$, and the remainder dominates). Thus:
\[
\int_{A_j} (\eta_0(x) - \eta_0(x_j))^2 dP \le P(A_j) \cdot C^2 (\delta_j)^{2\beta} \|\eta_0\|_{H^\beta}^2.
\]
Summing over $j$ and using $\delta_j = O(s^{-1/d})$ and $\sum_j P(A_j) = 1$:
\[
\|f_s - \eta_0\|_{L^2(P)}^2 \le C \sum_j P(A_j) \cdot s^{-2\beta/d} = C s^{-2\beta/d}.
\]
Thus $\inf_{f \in \cT_s} \|f - \eta_0\|_{L^2(P)}^2 \le C s^{-2\beta/d}$. By the loss--norm link, $\inf_{f \in \cT_s} [R(f) - R(\eta_0)] \le C s^{-2\beta/d}$ for squared loss and other losses satisfying $\phi(\delta) \lesssim \delta$. For further discussion of piecewise-constant approximation to Hölder functions, see Devore \& Lorentz (1993), Chapter~6 (approximation by piecewise polynomials), or Györfi et al.\ (2002), Theorem~3.2 (rates for $H^\beta$ in nonparametric regression).
\end{proof}

\begin{lemma}[Complexity of tree class]
\label{lem:complexity}
The covering number satisfies $\log N(\epsilon, \cT_s, L^2(P)) \lesssim s \log(1/\epsilon)$, or equivalently the VC dimension of $\cT_s$ is of order $s \log s$.
\end{lemma}
\begin{proof}
We define the relevant notions, bound the VC (pseudo-) dimension of $\cT_s$, and deduce the covering number bound.

\noindent\textbf{Definitions.}
Assume functions in $\cT_s$ are uniformly bounded (e.g.\ in $[0,1]$), which holds for propensity and bounded outcomes. The \emph{covering number} $N(\epsilon, \cT_s, L^2(P))$ is the minimal number of closed $L^2(P)$-balls of radius $\epsilon$ whose union contains $\cT_s$. The \emph{VC dimension} of a class of sets is the largest $n$ such that some set of $n$ points is shattered. For real-valued functions, the \emph{pseudo-dimension} (VC subgraph dimension) is the VC dimension of the class of subgraphs. A standard fact (e.g.\ van der Vaart and Wellner, 1996, Theorem 2.6.7) is that for a uniformly bounded function class with VC subgraph dimension at most $d$, $\log N(\epsilon, \cT_s, L^2(P)) \le C d \log(1/\epsilon)$ for $\epsilon$ small enough.

\noindent\textbf{VC bound.}
Each $f \in \cT_s$ is piecewise constant on an axis-aligned partition with at most $s$ cells; that is, $f = \sum_{j=1}^s a_j \mathbf{1}_{A_j}$ for disjoint axis-aligned cells $A_j$ and constants $a_j \in [0,1]$. The VC (pseudo-) dimension of this class is $O(s \log s)$: the $\log s$ factor arises from the combinatorial structure of binary trees, since a tree with $s$ leaves has $\approx 4^s$ distinct topologies (Catalan numbers), and each of the $s-1$ internal nodes requires choosing a split coordinate and threshold, yielding $O(s \log s)$ effective degrees of freedom. The class $\cT_s$ is contained in the set of linear combinations of $s$ such indicators with coefficients in $[0,1]$, so its pseudo-dimension is $O(s \log s)$. See Appendix~\ref{app:vc-dimension} for a complete self-contained proof, or Bartlett, Jordan, \& McAuliffe (2006, Lemma~9) for the explicit bound $V_{\text{sub}}(\cT_s) \le 4ds \log_2(3s)$ where $d$ is the number of (binary) features.

\noindent\textbf{Covering number.}
By the standard entropy bound (van der Vaart \& Wellner, 1996, Theorem~2.6.7), for a uniformly bounded function class with VC subgraph dimension at most $V$, $\log N(\epsilon, \cT_s, L^2(P)) \le K V (1 + \log(M/\epsilon))$ for a universal constant $K$ and $M = \sup_{f \in \cT_s} \|f\|_\infty$. With $V = O(s \log s)$ and $M = 1$ (functions in $[0,1]$), we obtain:
\[
\log N(\epsilon, \cT_s, L^2(P)) \le C s \log s \cdot (1 + \log(1/\epsilon)) = O(s \log s \cdot \log(1/\epsilon)).
\]
Thus the VC (pseudo-) dimension of $\cT_s$ is of order $s \log s$, and the covering number bound above suffices for the oracle inequality in Lemma~\ref{lem:oracle} (the complexity term $s_n \log n / n$ is consistent with $\log N \lesssim s \log s \cdot \log(1/\epsilon)$ when $\epsilon \sim 1/\sqrt{n}$, giving $\log N \lesssim s \log n$ for $s \ge 2$). In Lemma~\ref{lem:oracle} we apply this with $s = O(s_n)$ (since $\hat\eta \in \cT_{O_p(s_n)}$ and $f_n \in \cT_{s_n}$ by Lemma~\ref{lem:hats}).
\end{proof}

\begin{lemma}[Number of leaves of the penalized minimizer]
\label{lem:hats}
Let $\hat{s} := \#\text{leaves of }\hat\eta$. Under $\lambda_n \asymp (\log n)/n$ and $s_n \asymp n^{d/(2\beta+d)}$, the penalized minimizer $\hat\eta$ in \eqref{eq:perm} satisfies $\hat{s} = O_p(s_n)$.
\end{lemma}
\begin{proof}[Proof idea]
Take $f_n \in \cT_{s_n}$ with $R(f_n) - R(\eta_0) \le \inf_{f \in \cT_{s_n}}[R(f)-R(\eta_0)] + o(1)$. By definition of $\hat\eta$, $R_n(\hat\eta) + \lambda_n \hat{s} \le R_n(f_n) + \lambda_n s_n$, so $\lambda_n \hat{s} \le R_n(f_n) - R_n(\hat\eta) + \lambda_n s_n \le R_n(f_n) + \lambda_n s_n$. Boundedness of $R(\cdot)$ and the usual empirical-process bound give $R_n(f_n) \le R(f_n) + O_p(\sqrt{s_n \log n / n}) = O(1) + O_p(\sqrt{s_n \log n / n})$. Thus $\hat{s} \le s_n + (R_n(f_n) - R_n(\hat\eta))/\lambda_n \le s_n + O_p(1/\lambda_n) + O_p(\sqrt{s_n \log n / n}/\lambda_n)$. With $\lambda_n \asymp (\log n)/n$, the second term is $O_p(n/\log n)$ and the third is $O_p(n \sqrt{s_n \log n / n} / \log n) = O_p(\sqrt{n s_n / \log n})$. For $s_n \asymp n^{d/(2\beta+d)}$, $\sqrt{n s_n}/\sqrt{\log n}$ is $o(s_n)$ when $d/(2\beta+d) < 1$ (which holds for $\beta > 0$), so $\hat{s} \le s_n + o_p(s_n) = O_p(s_n)$.
\end{proof}

The following lemma verifies that the conditions of \citet{xuStatisticalOptimalityOptimal2026}, Theorem 3.1, hold in our setting, enabling us to apply their oracle inequality for penalized ERM.

\begin{lemma}[Verification of Xu et al.\ (2026) conditions]
\label{lem:xu-verification}
The setting of Theorem~\ref{thm:nuisance-rate} satisfies the conditions of \citet{xuStatisticalOptimalityOptimal2026}, Theorem 3.1:

\begin{enumerate}
\item \textbf{Bounded outcomes and predictions:} For squared loss with $Y \in [0,1]$, both outcomes and tree predictions are bounded. For cross-entropy with $Y \in \{0,1\}$ and $\eta_0(x) \in [c, 1-c]$, the loss $L(y,p) = -y \log p - (1-y) \log(1-p)$ is bounded for $p \in [c,1-c]$. We take $M = 1$ (maximum prediction value).

\item \textbf{Sub-Gaussian or bounded noise:} For squared loss, the noise $\varepsilon = Y - \eta_0(X)$ satisfies $|\varepsilon| \le 2M = 2$ (since $Y, \eta_0 \in [0,1]$), hence is sub-Gaussian with parameter $K = 2$. For cross-entropy with binary $Y$, the noise structure is implicit in the Bregman divergence (see Appendix~\ref{app:loss-norm}), and boundedness ensures finite variance.

\item \textbf{Penalty choice:} \citet{xuStatisticalOptimalityOptimal2026}, equation (7), requires $\lambda \ge C(M+K)(\log(nd)+u)/(\delta n)$ for appropriate constants. Our choice $\lambda_n \asymp (\log n)/n$ matches this requirement with $\delta \asymp \log n$, $u = 0$, and $C$ absorbing $(M+K)$ (both $O(1)$). This gives $\lambda_n \asymp (\log n)/n$ as stated.

\item \textbf{Sieve condition:} The oracle inequality requires $\delta \cdot s_n = o(n)$ (see Xu et al., Remark 3.2). With $\delta \asymp \log n$ and $s_n \asymp n^{d/(2\beta+d)}$, we have $\delta \cdot s_n \asymp n^{d/(2\beta+d)} \log n = o(n)$ since $d/(2\beta+d) < 1$ (equivalent to $\beta > 0$).

\item \textbf{Entropy condition:} The covering number bound $\log N(\epsilon, \cT_s, L^2(P)) \lesssim s \log s \cdot \log(1/\epsilon)$ required by Xu et al.\ is established in Lemma~\ref{lem:complexity} for axis-aligned decision trees.

\item \textbf{Cross-fitting adjustment:} Each fold trains on $n_{-k} = (K-1)n/K$ observations. Since $K$ is fixed, $(K-1)/K$ is a constant (e.g., $4/5$ for $K=5$), so the rate $n_{-k}^{-\beta/(2\beta+d)} \asymp n^{-\beta/(2\beta+d)}$ applies uniformly across folds, with the implicit constant depending on $K$.
\end{enumerate}
\end{lemma}

\begin{proof}
Conditions 1, 2, 4, and 5 follow directly from our setup (bounded outcomes for squared loss and cross-entropy, sieve growth rate, entropy bound from Lemma~\ref{lem:complexity}). Condition 3 holds by matching Xu et al.'s penalty requirement with our choice $\lambda_n \asymp (\log n)/n$. Condition 6 accounts for the reduced sample size per fold in cross-fitting, which affects only the constant in the rate, not the polynomial order.
\end{proof}

The following oracle inequality is a direct application of \citet{xuStatisticalOptimalityOptimal2026}, Theorem 3.1, whose conditions are verified in Lemma~\ref{lem:xu-verification}.

\begin{lemma}[Sharp oracle inequality for penalized ERM]
\label{lem:oracle}
For $\hat\eta$ the penalized ERM~\eqref{eq:perm} (minimize $R_n(f) + \lambda_n \cdot (\#\text{leaves of } f)$ over $\cT_{M_n}$) with $\lambda_n \asymp (\log n)/n$ and $s_n$ as above,
\[
R(\hat\eta) - R(\eta_0) \le C \Bigl[ \inf_{f \in \cT_{s_n}} \bigl\{ R(f) - R(\eta_0) \bigr\} + \frac{s_n \log n}{n} \Bigr] + o_p(1).
\]
\end{lemma}

\begin{remark}[Approximation-estimation decomposition]
The oracle inequality \eqref{lem:oracle} decomposes the excess risk into two components: (1) the \textbf{approximation error} $\inf_{f \in \cT_{s_n}} [R(f) - R(\eta_0)]$, which quantifies how well trees with at most $s_n$ leaves can approximate the true function $\eta_0$; and (2) the \textbf{estimation error} $s_n \log n / n$, which reflects the complexity penalty for finite-sample estimation. The optimal complexity $s_n$ balances these two errors. This is the classical bias-variance trade-off for nonparametric estimation \citep{xuStatisticalOptimalityOptimal2026}.
\end{remark}

\begin{proof}
This follows from \citet{xuStatisticalOptimalityOptimal2026}, Theorem 3.1, applied to our penalized ERM setting with $\lambda \asymp (\log n)/n$. The oracle inequality bounds the excess risk of the penalized ERM estimator in terms of the approximation error (best tree in $\cT_{s_n}$) and the estimation error ($s_n \log n / n$), which is the classical bias-variance decomposition for penalized empirical risk minimization. For complete details, see \citet{xuStatisticalOptimalityOptimal2026}, Section 3.

The key steps: (1) decompose excess risk into approximation (how well the sieve $\cT_{s_n}$ approximates $\eta_0$) and estimation (empirical process fluctuation); (2) control the empirical process using Rademacher complexity and uniform concentration; and (3) show that the penalty $\lambda_n s_n$ balances these errors when $\lambda_n \asymp (\log n)/n$. \citet{xuStatisticalOptimalityOptimal2026} establish sharp bounds via empirically localized Rademacher complexity, yielding the stated rate. By Lemma~\ref{lem:hats}, the penalized minimizer $\hat\eta$ has $O_p(s_n)$ leaves, so the oracle inequality applies with $s_n$ as the complexity index.

\noindent\textbf{Remark: $\varepsilon_n$-optimal minimizers.}
We say $\hat\eta$ is \emph{$\varepsilon_n$-optimal} if its penalized empirical risk is at most $\inf_{f \in \cT_{M_n}}(\cdots) + \varepsilon_n$. Then $R_n(\hat\eta) - R_n(f_n) \le \lambda_n s_n + \varepsilon_n$; the same oracle inequality holds provided $\varepsilon_n = o(s_n \log n / n)$ (e.g., $\varepsilon_n = o(1/n)$), with $\varepsilon_n$ absorbed into the $o_p(1)$ term. GOSDT-style solvers with an appropriate tolerance produce such an $\hat\eta$.
\end{proof}

\begin{lemma}[Optimal choice of $s_n$]
\label{lem:optimal-sn}
With $s_n$ as the rate index (the penalized minimizer has $\hat{s} = O_p(s_n)$ leaves by Lemma~\ref{lem:hats}), balance the approximation term $s_n^{-2\beta/d}$ and the estimation term $s_n (\log n)/n$. Setting $s_n \asymp n^{d/(2\beta+d)}$ yields excess risk $R(\hat\eta) - R(\eta_0) = O_p(n^{-2\beta/(2\beta+d)})$. By the loss--norm link, $\|\hat\eta - \eta_0\|_{L^2(P)}^2 \le \phi(O_p(n^{-2\beta/(2\beta+d)})) = O_p(n^{-2\beta/(2\beta+d)})$ (e.g., if $\phi(\delta) \lesssim \delta$), so
\[
\|\hat\eta - \eta_0\|_{L^2(P)} = O_p\bigl( n^{-\beta/(2\beta+d)} \bigr).
\]
This is the classical minimax rate for $H^\beta([0,1]^d)$.
\end{lemma}
\begin{proof}[Proof idea]
Set $s_n^{-2\beta/d} \asymp s_n (\log n)/n$ and solve for $s_n$; substitute into the oracle inequality to get the excess-risk rate; apply the loss--norm link to obtain the $L^2(P)$ rate.
\end{proof}

The following theorem combines the oracle inequality framework from \citet{xuStatisticalOptimalityOptimal2026} with the loss-norm link (Lemma~\ref{lem:loss-norm}) and the semiparametric rate condition to establish the convergence rate of tree-based nuisance estimators for causal inference.

\begin{theorem}[Nuisance rate]
\label{thm:nuisance-rate}
Suppose $\eta_0 \in H^\beta([0,1]^d)$, $X \in [0,1]^d$, the loss is either (i) squared error $L(y,f) = (y-f)^2$ with $Y \in [0,1]$, or (ii) cross-entropy $L(y,p) = -y \log p - (1-y) \log(1-p)$ with $Y \in \{0,1\}$ and $\eta_0(x) \in [c, 1-c]$ for some $c > 0$. Let $\hat\eta$ be the penalized ERM~\eqref{eq:perm} (loss plus $\lambda_n \times \#\text{leaves}$) over $\cT_{M_n}$ with $\lambda_n \asymp (\log n)/n$ and $s_n \asymp n^{d/(2\beta+d)}$. Then $\|\hat\eta - \eta_0\|_{L^2(P)} = O_p(n^{-\beta/(2\beta+d)})$. If $\beta/(2\beta+d) > 1/4$, then this rate is $o_p(n^{-1/4})$.
\end{theorem}
\begin{proof}
This theorem combines three components: (1) the oracle inequality from \citet{xuStatisticalOptimalityOptimal2026}, Theorem 3.1, which establishes the excess risk bound for penalized ERM trees; (2) the approximation rate for Hölder-$\beta$ functions (Lemma~\ref{lem:approx}); and (3) the loss-norm link (Lemma~\ref{lem:loss-norm}). We establish the result through the following steps.

We combine Lemmas~\ref{lem:approx}--\ref{lem:optimal-sn} to obtain the stated rate. The proof proceeds in five steps: apply the oracle inequality, substitute the optimal $s_n$, derive the excess risk rate, apply the loss--norm link, and take the square root.

\noindent\textbf{Step 1: Apply the oracle inequality (Lemma~\ref{lem:oracle}).}
By Lemma~\ref{lem:oracle}, for $\hat\eta$ the penalized ERM with $\lambda_n \asymp (\log n)/n$:
\[
R(\hat\eta) - R(\eta_0) \le C \Bigl[ \inf_{f \in \cT_{s_n}} \bigl\{ R(f) - R(\eta_0) \bigr\} + \frac{s_n \log n}{n} \Bigr] + o_p(1).
\]

\noindent\textbf{Step 2: Apply the approximation bound (Lemma~\ref{lem:approx}).}
By Lemma~\ref{lem:approx}, for $\eta_0 \in H^\beta([0,1]^d)$:
\[
\inf_{f \in \cT_{s_n}} \bigl[ R(f) - R(\eta_0) \bigr] \le C' s_n^{-2\beta/d}.
\]

Substituting into Step~1:
\[
R(\hat\eta) - R(\eta_0) \le C \Bigl( C' s_n^{-2\beta/d} + \frac{s_n \log n}{n} \Bigr) + o_p(1).
\]

\noindent\textbf{Step 3: Choose optimal $s_n$ (Lemma~\ref{lem:optimal-sn}).}
By Lemma~\ref{lem:optimal-sn}, we balance the approximation term $s_n^{-2\beta/d}$ and the estimation term $s_n \log n / n$ by setting:
\[
s_n^{-2\beta/d} \asymp \frac{s_n \log n}{n}.
\]

Solving for $s_n$:
\[
s_n^{1 + 2\beta/d} \asymp \frac{n}{\log n} \implies s_n \asymp \Bigl( \frac{n}{\log n} \Bigr)^{d/(d + 2\beta)}.
\]

For asymptotic analysis, we absorb polylogarithmic factors into the $\asymp$ symbol, noting that they do not affect the polynomial rate:
\[
s_n \asymp n^{d/(2\beta + d)}.
\]

With this choice, both terms in the oracle inequality have the same order:
\[
s_n^{-2\beta/d} \asymp \frac{s_n \log n}{n} \asymp n^{-2\beta/(2\beta+d)}.
\]

(To see this: $s_n^{-2\beta/d} = (n^{d/(2\beta+d)})^{-2\beta/d} = n^{-2\beta/(2\beta+d)}$.)

Thus:
\[
R(\hat\eta) - R(\eta_0) \le C'' n^{-2\beta/(2\beta+d)} + o_p(1) = O_p\bigl( n^{-2\beta/(2\beta+d)} \bigr).
\]

\noindent\textbf{Step 4: Apply the loss--norm link.}
By the loss--norm link assumption (stated in the setup, line~187, and proved for squared loss and cross-entropy in Appendix~\ref{app:loss-norm}), there exists $\phi(\delta) \to 0$ as $\delta \to 0$ such that:
\[
R(f) - R(\eta_0) \le \delta \implies \|f - \eta_0\|_{L^2(P)}^2 \le \phi(\delta).
\]

For squared loss, $\phi(\delta) = \delta$ (identity). For cross-entropy with bounded probabilities, $\phi(\delta) = C\delta$ for a constant $C$ (Lemma~\ref{lem:ce-loss-norm}). In both cases, $\phi(\delta) \lesssim \delta$.

From Step~3, $R(\hat\eta) - R(\eta_0) = O_p(n^{-2\beta/(2\beta+d)})$. Applying the loss--norm link:
\[
\|\hat\eta - \eta_0\|_{L^2(P)}^2 \le \phi\bigl\{ O_p(n^{-2\beta/(2\beta+d)}) \bigr\} = O_p(n^{-2\beta/(2\beta+d)}),
\]
where the last equality uses $\phi(\delta) \lesssim \delta$.

\noindent\textbf{Step 5: Take the square root.}
Since $\|\hat\eta - \eta_0\|_{L^2(P)}^2 = O_p(n^{-2\beta/(2\beta+d)})$, taking square roots:
\[
\|\hat\eta - \eta_0\|_{L^2(P)} = O_p\bigl( n^{-\beta/(2\beta+d)} \bigr).
\]

This is the classical minimax rate for estimating $\eta_0 \in H^\beta([0,1]^d)$ in $L^2$ norm, achieved by the tree estimator with optimal regularization $s_n \asymp n^{d/(2\beta+d)}$.

\noindent\textbf{Sufficiency for valid inference (second part of theorem).}
The theorem also states that if $\beta/(2\beta+d) > 1/4$, then the rate $n^{-\beta/(2\beta+d)}$ is faster than $n^{-1/4}$, i.e., $\|\hat\eta - \eta_0\|_{L^2(P)} = o_p(n^{-1/4})$.

To verify: $\beta/(2\beta+d) > 1/4$ is equivalent to $4\beta > 2\beta + d$, which simplifies to $2\beta > d$, or $\beta > d/2$. This is the smoothness condition stated in the theorem.

When $\beta > d/2$, the convergence rate $n^{-\beta/(2\beta+d)} = o(n^{-1/4})$, which by semiparametric theory (Chernozhukov et al., 2018) is sufficient for $\sqrt{n}$-consistent and asymptotically normal estimation of the causal parameter (Corollary~\ref{cor:dml}). This completes the proof.
\end{proof}

\begin{corollary}[Valid inference with tree nuisances]
\label{cor:dml}
If each nuisance estimator satisfies $\|\hat\eta - \eta_0\|_{L^2(P)} = o_p(n^{-1/4})$ in the norms that appear in the orthogonal score expansion, then under identification (overlap, bounded moments) and $K$-fold cross-fitting, the EIF-based estimator satisfies $\sqrt{n}(\hat\theta - \theta_0) \leadsto N(0, \sigma^2)$, where $\sigma^2$ is the variance of the efficient influence function.
\end{corollary}

\begin{proof}
We invoke the general semiparametric theory of \citet{chernozhukov2018} and verify that their conditions hold for tree-based nuisance estimators.

\noindent\textbf{Chernozhukov et al.\ (2018) framework.}
By Theorem~\ref{thm:nuisance-rate}, each tree nuisance estimator $\hat\eta^{(-k)}$ (trained on data not in fold $k$) achieves $\|\hat\eta^{(-k)} - \eta_0\|_{L^2(P)} = O_p(n^{-\beta/(2\beta+d)}) = o_p(n^{-1/4})$ when $\beta > d/2$. This applies componentwise to propensity $\hat{e}^{(-k)}$ and outcome regression $\hat{\mu}_0^{(-k)}$.

Chernozhukov et al.\ (2018, Theorem~3.1) establish that under:
\begin{enumerate}[label=(\roman*)]
\item \textbf{Neyman orthogonality:} The moment function $\psi(O; \theta, \eta)$ satisfies $\partial_\eta \E[\psi(O; \theta_0, \eta_0)] = 0$ (score is orthogonal to nuisance perturbations at $(\theta_0, \eta_0)$).
\item \textbf{Nuisance rate condition:} The second-order remainder term is $o_p(n^{-1/2})$.
\item \textbf{Identification:} $\E[\psi(O; \theta_0, \eta_0)] = 0$ uniquely identifies $\theta_0$, with non-singular Jacobian $J = \partial_\theta \E[\psi(O; \theta_0, \eta_0)]$.
\item \textbf{Regularity:} Bounded moments, overlap, and cross-fitting with $K$ fixed.
\end{enumerate}
the estimator $\hat\theta$ defined by $(1/n)\sum_i \psi(O_i; \hat\theta, \hat\eta^{(-k(i))}) = 0$ satisfies $\sqrt{n}(\hat\theta - \theta_0) \leadsto N(0, J^{-1} \Omega J^{-1})$ where $\Omega = \text{Var}[\psi(O; \theta_0, \eta_0)]$.

We now verify each condition for the ATT with tree nuisances.

\noindent\textbf{Condition (i): Neyman orthogonality.}
For ATT, the orthogonal score is (e.g., Hahn 1998; Chernozhukov et al.\ 2018, Example~2):
\[
\psi(O; \theta, \eta) = \frac{A}{\pi}\bigl\{ Y - \mu_0(X) - \theta \bigr\} + \frac{1-A}{1-\pi} \, \frac{e(X)}{1 - e(X)} \bigl\{ Y - \mu_0(X) \bigr\},
\]
where $\eta = (e, \mu_0)$, $\pi = \Prob(A=1)$, $e(X) = \Prob(A=1|X)$, and $\mu_0(X) = \E[Y|A=0,X]$. This score satisfies $\partial_\eta \E[\psi(O; \theta_0, \eta_0)] = 0$ at $(\theta_0, \eta_0)$ by construction (this is the defining property of Neyman-orthogonal scores). See Chernozhukov et al.\ (2018), Section~3.2 for the derivation. $\checkmark$

\noindent\textbf{Condition (ii): Second-order remainder is $o_p(n^{-1/2})$.}
By Chernozhukov et al.\ (2018, Lemma~3.1 and their equation (3.11)), the second-order remainder for the orthogonal score is bounded by:
\[
|\text{Remainder}| \le C \cdot \|\hat{e} - e_0\|_{L^2(P_1)} \cdot \|\hat{\mu}_0 - \mu_{0,0}\|_{L^2(P_1)} + \text{similar cross-terms},
\]
where $P_1$ is the marginal distribution of $(X, A)$ (or $P$ if $X, A$ are independent of fold membership). Under overlap ($0 < c \le e_0(X) \le 1-c < 1$), the weights $1/e(X)$ and $1/(1-e(X))$ are bounded, so the relevant norms are the standard $L^2(P)$ norms (Assumption~3.2 in Chernozhukov et al.\ 2018).

For our tree estimators, $\|\hat{e}^{(-k)} - e_0\|_{L^2(P)} = o_p(n^{-1/4})$ and $\|\hat{\mu}_0^{(-k)} - \mu_{0,0}\|_{L^2(P)} = o_p(n^{-1/4})$ (by Theorem~\ref{thm:nuisance-rate} with $\beta > d/2$). Thus each product term satisfies:
\[
\|\hat{e}^{(-k)} - e_0\|_{L^2(P)} \cdot \|\hat{\mu}_0^{(-k)} - \mu_{0,0}\|_{L^2(P)} = o_p(n^{-1/4}) \cdot o_p(n^{-1/4}) = o_p(n^{-1/2}).
\]
Hence $|\text{Remainder}| = o_p(n^{-1/2})$. $\checkmark$

\noindent\textbf{Condition (iii): Identification.}
For ATT, the moment condition $\E[\psi(O; \theta_0, \eta_0)] = 0$ identifies $\theta_0 = \E[Y(1) - Y(0) | A = 1]$ (the average treatment effect on the treated) under unconfoundedness and overlap. This is standard in the causal inference literature (Hahn, 1998; Abadie, 2005; Imbens \& Wooldridge, 2009). The Jacobian $J = \E[\partial_\theta \psi(O; \theta_0, \eta_0)] = -\E[A/\pi]$ = $-1$ is non-zero. $\checkmark$

\noindent\textbf{Condition (iv): Regularity.}
\begin{itemize}
\item \textbf{Bounded moments:} $Y$ bounded (or finite variance) by assumption.
\item \textbf{Overlap:} $0 < c \le e_0(X) \le 1-c < 1$ by assumption (required for cross-entropy loss-norm link, Appendix~\ref{app:loss-norm}).
\item \textbf{Cross-fitting:} We use $K$-fold cross-fitting with $K$ fixed (e.g., $K = 2, 5, 10$), as described in the main text.
\end{itemize}
All conditions are satisfied. $\checkmark$

\noindent\textbf{Conclusion.}
All conditions of Chernozhukov et al.\ (2018, Theorem~3.1) are verified. Thus:
\[
\sqrt{n}(\hat\theta - \theta_0) \leadsto N(0, \sigma^2),
\]
where $\sigma^2 = J^{-1} \Omega J^{-1}$ with $\Omega = \text{Var}[\psi(O; \theta_0, \eta_0)]$ (the variance of the influence function). This can be estimated consistently using sample splitting (Chernozhukov et al., 2018, Section~4).
\end{proof}

\subsection{EIF-based inference with alternative nuisances (general theorem)}
\label{app:alt-nuisances}

The following result is of independent interest for cross-fitted inference: it states that any fold-specific nuisance estimators that are ``close'' to the true nuisances in the sense that controls the remainder term still yield $\sqrt{n}$-consistent, asymptotically normal inference. No structure (e.g., trees) is assumed; the result is generic.

\begin{definition}[$o_p(n^{-1/4})$-close nuisances]
\label{def:close}
For each fold $k \in \{1,\ldots,K\}$, let $\tilde{\eta}^{(-k)} = (\tilde{e}^{(-k)}, \tilde{m}_0^{(-k)}, \tilde{m}_1^{(-k)})$ be a nuisance estimator trained on data not in fold $k$. We say the $\tilde{\eta}^{(-k)}$ are \emph{$o_p(n^{-1/4})$-close} to $\eta_0$ if, for each $k$, we have $\|\tilde{e}^{(-k)} - e_0\|_{L^2(P)} = o_p(n^{-1/4})$, $\|\tilde{m}_0^{(-k)} - m_{0,0}\|_{L^2(P)} = o_p(n^{-1/4})$, and $\|\tilde{m}_1^{(-k)} - m_{1,0}\|_{L^2(P)} = o_p(n^{-1/4})$. Under the overlap condition ($0 < e_0(X) < 1$ bounded away from 0 and 1), these componentwise $L^2(P)$ norms are the relevant norms for the ATT score expansion; equivalently, $\max_{k=1,\ldots,K} \|\tilde{\eta}^{(-k)} - \eta_0\|_{L^2(P)} = o_p(n^{-1/4})$ when $\|\tilde{\eta}^{(-k)} - \eta_0\|_{L^2(P)}$ denotes the maximum of the three component norms.
\end{definition}

\begin{theorem}[Valid inference with alternative nuisances]
\label{thm:dml-alt}
Assume $K$ is fixed. Under the same identification, overlap, and moment conditions as in \citet{chernozhukov2018} (unconfoundedness, overlap, bounded outcomes and propensity), suppose that for each fold $k$, $\tilde{\eta}^{(-k)}$ is trained on data not in fold $k$ and is $o_p(n^{-1/4})$-close to $\eta_0$ in the sense of Definition~\ref{def:close}. Let $\hat\theta$ be the EIF-based estimator defined by
\[
\frac{1}{n} \sum_{i=1}^n \psi\bigl(O_i; \hat\theta, \tilde{\eta}^{(-k(i))}\bigr) = 0.
\]
Then $\sqrt{n}(\hat\theta - \theta_0) \leadsto N(0, \sigma^2)$, where $\sigma^2$ is the variance of the influence function $\psi(O; \theta_0, \eta_0)$.
\end{theorem}
\begin{proof}
By \citet{chernozhukov2018}, the only nuisance-dependent condition for $\sqrt{n}$-consistency and asymptotic normality is that the second-order remainder in the linearization of the moment condition be $o_p(n^{-1/2})$. For the ATT orthogonal score \eqref{eq:score}, the remainder is bounded in absolute value by a constant times products of the form $\|\tilde{e}^{(-k)} - e_0\|_{L^2(P)} \|\tilde{m}_a^{(-k)} - m_{a,0}\|_{L^2(P)}$ (and similar cross-terms), possibly with overlap-weighted norms; see \citet{chernozhukov2018}, Section~3.2 and their Assumption 3.2. Under overlap, the weights are bounded, so componentwise $\|\tilde{\eta}^{(-k)} - \eta_0\|_{L^2(P)} = o_p(n^{-1/4})$ implies that each such product is $o_p(n^{-1/2})$. Hence the remainder is $o_p(n^{-1/2})$. The remaining conditions in \citet{chernozhukov2018} (identification, cross-fitting with $K$ fixed, bounded moments) do not depend on the choice of nuisance estimator. Thus their conclusion applies with $\tilde{\eta}^{(-k)}$ in place of $\hat\eta^{(-k)}$.
\end{proof}

\subsection{Rashomon sets and intersection: theory}
\label{app:rashomon}

We give formal definitions of the Rashomon set for each cross-fitting fold and of the intersection across folds, along with existence results and finite-sample bounds. These are used in Section~\ref{sec:rashomon}. For a fixed nuisance (e.g., propensity $e$ or outcome regression $m_0$), the training data for fold $k$ is $\mathcal{D}^{(-k)}$ (all observations not in fold $k$); $n_{-k} = |\mathcal{D}^{(-k)}|$. Let $R_n^{(-k)}(f) = (1/n_{-k}) \sum_{i \notin \text{fold } k} L(Y_i, f(X_i))$ denote the empirical risk on $\mathcal{D}^{(-k)}$ for the relevant loss $L$, and let $\text{best}^{(-k)} = \inf_{f \in \cT_{M_n}} [ R_n^{(-k)}(f) + \lambda_n \cdot (\#\text{leaves of } f) ]$ be the minimal penalized risk on that fold.

\subsubsection{Non-empty Rashomon set per fold}

\begin{lemma}[Non-empty Rashomon set per fold]
\label{lem:rashomon-nonempty}
For any fold $k$ and any $\varepsilon \ge 0$, the Rashomon set $R_k(\varepsilon)$ is non-empty: it contains at least the minimizer $\hat\eta^{(-k)}$ that achieves $\text{best}^{(-k)}$.
\end{lemma}
\begin{proof}
By definition, $\text{best}^{(-k)} = \inf_{f \in \cT_{M_n}} [ R_n^{(-k)}(f) + \lambda_n \cdot (\#\text{leaves of } f) ]$. For any $\varepsilon \ge 0$, the minimizer $\hat\eta^{(-k)}$ satisfies $R_n^{(-k)}(\hat\eta^{(-k)}) + \lambda_n \cdot (\#\text{leaves of } \hat\eta^{(-k)}) = \text{best}^{(-k)} \le (1+\varepsilon) \cdot \text{best}^{(-k)}$, so $\hat\eta^{(-k)} \in R_k(\varepsilon)$.
\end{proof}

\subsubsection{Intersection existence under stability}

\begin{theorem}[Intersection existence under stability]
\label{thm:intersection-existence}
Let $\varepsilon_n \to 0$ with $\varepsilon_n = o(n^{-1/2})$ as required for valid inference. Suppose the following \textbf{stability condition} holds:
\begin{equation}
\label{eq:stability}
\text{(A4) There exists a tree structure } \tau^* \text{ with } \#\text{leaves}(\tau^*) \le s_n \text{ such that, when refit on } \mathcal{D}^{(-k)} \text{ for each fold } k,
\end{equation}
the resulting tree $f_k^*$ (with structure $\tau^*$ and leaf values optimized on $\mathcal{D}^{(-k)}$) satisfies
\[
R_n^{(-k)}(f_k^*) + \lambda_n \cdot (\#\text{leaves of } f_k^*) \le (1+\varepsilon_n) \cdot \text{best}^{(-k)}, \quad \text{for all } k = 1, \ldots, K.
\]
Then the intersection $\bigcap_{k=1}^K R_k(\varepsilon_n)$ is non-empty: it contains the structure $\tau^*$.
\end{theorem}

\begin{proof}
By assumption (A4), there exists a structure $\tau^*$ such that for each fold $k$, the refit $f_k^*$ on $\mathcal{D}^{(-k)}$ belongs to $R_k(\varepsilon_n)$. By Definition~\ref{def:intersection}, a structure belongs to the intersection if and only if it appears in $R_k(\varepsilon_n)$ for every fold $k$ when refit per fold. Since $\tau^*$ satisfies this by assumption, $\tau^* \in \bigcap_{k=1}^K R_k(\varepsilon_n)$.
\end{proof}

\noindent\textbf{Interpretation of the stability assumption.}
The stability condition (A4) says: ``there exists a tree structure that is robustly good across folds.'' This is interpretable and leaves failure modes explicit. Specifically:
\begin{itemize}
\item \textbf{When (A4) is plausible:} If the true nuisance $\eta_0$ is well-approximated by a tree partition, and the data are homogeneous enough that the same partition works well on each fold's training set, then (A4) holds with high probability for large $n$. Smooth or piecewise-smooth functions, stable covariate distributions, and moderate $\varepsilon_n$ make this more likely.
\item \textbf{When (A4) may fail:} If the training sets across folds are too heterogeneous (e.g., small sample size, high-dimensional covariates, or rough functions requiring fold-specific partitions), no single structure may be $(1+\varepsilon_n)$-near-optimal on all folds. In such cases, the intersection can be empty.
\end{itemize}

We do not prove that (A4) holds in general; that would require uniform convergence over structures and fold-specific training sets, which is beyond the scope of the current work. Instead, we verify (A4) empirically in simulations (Section~\ref{sec:results}), where we observe the frequency with which the intersection is non-empty across different data-generating processes, sample sizes, and tolerance levels $\varepsilon_n$.

\subsubsection{When the intersection can be empty}

\begin{remark}[When the intersection can be empty]
\label{rem:intersection-empty}
The intersection $\bigcap_{k=1}^K R_k(\varepsilon_n)$ can be empty in the following scenarios:
\begin{enumerate}
\item \textbf{Very small tolerance $\varepsilon_n$:} If $\varepsilon_n$ is close to zero, the Rashomon sets per fold shrink to near-singletons (just the minimizer), and fold-specific minimizers may differ in structure. As $\varepsilon_n \to 0$, the probability that a common structure exists goes to zero unless the folds are nearly identical.
\item \textbf{Small sample size $n$:} For small $n$, fold-specific training sets may be too variable for a single structure to perform well on all folds. As $n \to \infty$ and $\varepsilon_n \to 0$ appropriately, the intersection becomes more likely to be non-empty if (A4) holds asymptotically.
\item \textbf{Heterogeneous covariate distributions across folds:} If the folds have systematically different covariate distributions (e.g., due to imbalance in sampling or strong time trends in longitudinal data), no single partition may fit all folds well.
\item \textbf{High-dimensional or rough nuisances:} Functions that require complex, fold-specific partitions (high $d$, low $\beta$, or many interactions) are less likely to have a single structure in the intersection.
\end{enumerate}
In practice, if the intersection is empty, the implementation can fall back to fold-specific trees (as in the \texttt{doubletree} package with \texttt{use\_rashomon = TRUE}). Empirically, we find that the intersection is non-empty in 100\% of replications for the data-generating processes tested (DGPs 1--4, Section~\ref{sec:results}), including the weak overlap stress test (DGP 4). However, \textbf{non-emptiness does not guarantee valid inference}: DGP 4 exhibits a non-empty intersection in all replications but doubletree-based inference shows substantial coverage degradation (53.5\% vs.\ nominal 95\%) due to extreme propensity score weights. This highlights the distinction between \emph{structural existence} (intersection non-empty, Assumption~A4 satisfied) and \emph{statistical adequacy} (valid asymptotic inference under finite-sample conditions). Performance quality depends on both the intersection existing and the nuisance approximation quality being adequate under the actual overlap and smoothness conditions.
\end{remark}

\subsubsection{Formal definitions}

\begin{definition}[Rashomon set per fold]
\label{def:rashomon-fold}
For $\varepsilon \ge 0$, the \emph{Rashomon set} for fold $k$ is
\[
R_k(\varepsilon) = \Bigl\{ f \in \cT_{M_n} : R_n^{(-k)}(f) + \lambda_n \cdot (\#\text{leaves of } f) \le (1+\varepsilon) \cdot \text{best}^{(-k)} \Bigr\}.
\]
Optionally one restricts to trees with $\#\text{leaves} \le s_n$ so that the set is compatible with the sieve. We say trees in $R_k(\varepsilon)$ are \emph{$(1+\varepsilon)$-near-optimal} on fold $k$.
\end{definition}

\begin{lemma}[Oracle inequality for near-minimizers]
\label{lem:oracle-near}
Under the same conditions as Lemma~\ref{lem:oracle} (with $R_n^{(-k)}$, $\mathcal{D}^{(-k)}$, and $\text{best}^{(-k)}$ in place of $R_n$, $n$, and the fold-$k$ minimizer). For any $f \in R_k(\varepsilon_n)$ with $\#\text{leaves}(f) \le s_n$ (or $\#\text{leaves}(f) = O_p(s_n)$ under the same $\lambda_n$ logic as Lemma~\ref{lem:hats}),
\[
R(f) - R(\eta_0) \le C \Bigl[ \varepsilon_n \cdot \text{best}^{(-k)} + \frac{s_n \log n}{n} + \inf_{f' \in \cT_{s_n}} \bigl\{ R(f') - R(\eta_0) \bigr\} \Bigr] + o_p(1).
\]
\end{lemma}
\begin{proof}
We prove the oracle inequality for any $f \in R_k(\varepsilon_n)$ with $\#\text{leaves}(f) \le s_n$.

\noindent\textbf{Step 1: Setup.}
Let $f^* \in \cT_{s_n}$ be a tree such that $R(f^*) - R(\eta_0) \le \inf_{f' \in \cT_{s_n}}[R(f') - R(\eta_0)] + o(1)$. Since $f \in R_k(\varepsilon_n)$, we have
\[
R_n^{(-k)}(f) + \lambda_n \cdot (\#\text{leaves of } f) \le (1+\varepsilon_n) \cdot \text{best}^{(-k)}.
\]
By definition, $\text{best}^{(-k)} = \inf_{g \in \cT_{M_n}} [ R_n^{(-k)}(g) + \lambda_n \cdot (\#\text{leaves of } g) ] \le R_n^{(-k)}(f^*) + \lambda_n \cdot s_n$. Thus
\[
R_n^{(-k)}(f) \le (1+\varepsilon_n) \cdot \text{best}^{(-k)} \le (1+\varepsilon_n) \cdot [ R_n^{(-k)}(f^*) + \lambda_n \cdot s_n ].
\]

\noindent\textbf{Step 2: Decomposition.}
Decompose the excess risk $R(f) - R(\eta_0)$ as
\begin{align*}
R(f) - R(\eta_0) &= \underbrace{\bigl\{ R(f) - R_n^{(-k)}(f) \bigr\}}_{\text{(I)}} + \underbrace{\bigl\{ R_n^{(-k)}(f) - R_n^{(-k)}(f^*) \bigr\}}_{\text{(II)}} \\
&\quad + \underbrace{\bigl\{ R_n^{(-k)}(f^*) - R(f^*) \bigr\}}_{\text{(III)}} + \underbrace{\bigl\{ R(f^*) - R(\eta_0) \bigr\}}_{\text{(IV)}}.
\end{align*}

\noindent\textbf{Step 3: Bound term (II).}
From Step 1,
\[
R_n^{(-k)}(f) - R_n^{(-k)}(f^*) \le (1+\varepsilon_n) \cdot [ R_n^{(-k)}(f^*) + \lambda_n \cdot s_n ] - R_n^{(-k)}(f^*) = \varepsilon_n \cdot R_n^{(-k)}(f^*) + (1+\varepsilon_n) \lambda_n \cdot s_n.
\]
Since $R_n^{(-k)}(f^*) \le R(f^*) + O_p(\sqrt{s_n \log n / n})$ (by empirical process bounds, see Step 4) and $R(f^*) = R(\eta_0) + O(s_n^{-2\beta/d})$, we have $R_n^{(-k)}(f^*) = O_p(1)$. Write $R_n^{(-k)}(f^*) \le \text{best}^{(-k)} + O_p(s_n \log n / n)$ (since $f^*$ is near-optimal). Thus
\[
\text{(II)} \le \varepsilon_n \cdot \text{best}^{(-k)} + O(s_n \log n / n).
\]

\noindent\textbf{Step 4: Bound terms (I) and (III).}
Both $f$ and $f^*$ belong to $\cT_{s_n}$ (or $\cT_{C s_n}$ with high probability). By the same empirical process argument as in Lemma~\ref{lem:oracle}, Step 3 (using Lemma~\ref{lem:complexity} for entropy and maximal inequality), we have
\[
R(f) - R_n^{(-k)}(f) = O_p\bigl( \sqrt{s_n \log n / n} \bigr), \quad R_n^{(-k)}(f^*) - R(f^*) = O_p\bigl( \sqrt{s_n \log n / n} \bigr).
\]
With $s_n = n^{d/(2\beta+d)}$, this is $O_p(n^{-\beta/(2\beta+d)})$. Squaring, the contribution to excess risk is $O_p(s_n \log n / n)$.

\noindent\textbf{Step 5: Bound term (IV).}
By Lemma~\ref{lem:approx}, $R(f^*) - R(\eta_0) \le \inf_{f' \in \cT_{s_n}}[R(f') - R(\eta_0)] + o(1)$.

\noindent\textbf{Step 6: Combine.}
Summing the bounds,
\begin{align*}
R(f) - R(\eta_0) &\le O_p\bigl( \sqrt{s_n \log n / n} \bigr) + \varepsilon_n \cdot \text{best}^{(-k)} + O(s_n \log n / n) \\
&\quad + O_p\bigl( \sqrt{s_n \log n / n} \bigr) + \inf_{f' \in \cT_{s_n}}[R(f') - R(\eta_0)] + o(1).
\end{align*}
All $O_p(\sqrt{s_n \log n / n})$ terms sum to $O_p(s_n \log n / n)$ after squaring. Taking $C$ large enough,
\[
R(f) - R(\eta_0) \le C \Bigl[ \varepsilon_n \cdot \text{best}^{(-k)} + \frac{s_n \log n}{n} + \inf_{f' \in \cT_{s_n}} \bigl\{ R(f') - R(\eta_0) \bigr\} \Bigr] + o_p(1).
\]
\end{proof}

\begin{definition}[Intersection of Rashomon sets (structure-based)]
\label{def:intersection}
The \emph{intersection} of the Rashomon sets across folds is the set of \emph{tree structures} (splits) such that for every fold $k$, when that structure is refit on $\mathcal{D}^{(-k)}$ (i.e., leaf values are optimized on data not in fold $k$), the resulting tree belongs to $R_k(\varepsilon)$. Here, \emph{refit} means: minimize $R_n^{(-k)}(f) + \lambda_n \cdot (\#\text{leaves of } f)$ over all trees $f$ with that same structure (same splits, variable leaf values). Equivalently, a structure is in the intersection if it appears in $R_k(\varepsilon)$ for each $k$ when each fold's trees are fitted on $\mathcal{D}^{(-k)}$.
\end{definition}

\subsubsection{Finite-sample bounds for intersection trees}

\begin{lemma}[Boundedness of $\text{best}^{(-k)}$]
\label{lem:best-bounded}
Under the conditions of Theorem~\ref{thm:nuisance-rate}, the minimal penalized risk $\text{best}^{(-k)} = \inf_{f \in \cT_{M_n}} [ R_n^{(-k)}(f) + \lambda_n \cdot (\#\text{leaves of } f) ]$ satisfies $\text{best}^{(-k)} = O_p(1)$ for each fold $k$.
\end{lemma}

\begin{proof}
Let $f^* \in \cT_{s_n}$ be a tree with $R(f^*) - R(\eta_0) \le \inf_{f \in \cT_{s_n}}[R(f) - R(\eta_0)] + o(1)$, so that $R(f^*) = R(\eta_0) + O(s_n^{-2\beta/d})$ by Lemma~\ref{lem:approx}. Then
\[
\text{best}^{(-k)} \le R_n^{(-k)}(f^*) + \lambda_n \cdot s_n.
\]
By standard empirical process bounds (e.g., Lemma~\ref{lem:complexity} and uniform convergence for $\cT_{s_n}$), $R_n^{(-k)}(f^*) - R(f^*) = O_p(\sqrt{s_n \log n / n})$. Thus
\[
\text{best}^{(-k)} \le R(f^*) + O_p(\sqrt{s_n \log n / n}) + \lambda_n \cdot s_n = R(\eta_0) + O(s_n^{-2\beta/d}) + O_p(\sqrt{s_n \log n / n}) + O(s_n \log n / n).
\]
With $s_n = n^{d/(2\beta+d)}$, all terms are $O_p(1)$ (or $o_p(1)$), so $\text{best}^{(-k)} = O_p(1)$.
\end{proof}

\begin{lemma}[Intersection tree bias bound]
\label{lem:intersection-bias}
Suppose the intersection $\bigcap_{k=1}^K R_k(\varepsilon_n)$ is non-empty, and let $\tilde{\eta}^{(-k)}$ be the refit of an intersection structure on fold $k$'s training data. Then, for each fold $k$,
\[
R(\tilde{\eta}^{(-k)}) - R(\eta_0) \le C \Bigl[ \varepsilon_n \cdot \text{best}^{(-k)} + \frac{s_n \log n}{n} + \inf_{f \in \cT_{s_n}} \bigl\{ R(f) - R(\eta_0) \bigr\} \Bigr] + o_p(1).
\]
Under the conditions of Theorem~\ref{thm:nuisance-rate} (with $s_n \asymp n^{d/(2\beta+d)}$ and $\varepsilon_n = o(n^{-1/2})$), this simplifies to
\[
R(\tilde{\eta}^{(-k)}) - R(\eta_0) = O_p\bigl( \varepsilon_n + n^{-2\beta/(2\beta+d)} \bigr) = o_p(n^{-1/2}),
\]
and by the loss--norm link,
\[
\|\tilde{\eta}^{(-k)} - \eta_0\|_{L^2(P)} = O_p\bigl( \sqrt{\varepsilon_n} + n^{-\beta/(2\beta+d)} \bigr) = o_p(n^{-1/4}).
\]
\end{lemma}

\begin{proof}
\noindent\textbf{Step 1: Apply Lemma~\ref{lem:oracle-near}.}
By assumption, the intersection structure, when refit on $\mathcal{D}^{(-k)}$, yields a tree $\tilde{\eta}^{(-k)} \in R_k(\varepsilon_n)$ with $\#\text{leaves}(\tilde{\eta}^{(-k)}) \le s_n$. Lemma~\ref{lem:oracle-near} gives
\[
R(\tilde{\eta}^{(-k)}) - R(\eta_0) \le C \Bigl[ \varepsilon_n \cdot \text{best}^{(-k)} + \frac{s_n \log n}{n} + \inf_{f \in \cT_{s_n}} \bigl\{ R(f) - R(\eta_0) \bigr\} \Bigr] + o_p(1).
\]

\noindent\textbf{Step 2: Use boundedness.}
By Lemma~\ref{lem:best-bounded}, $\text{best}^{(-k)} = O_p(1)$. With $\varepsilon_n = o(n^{-1/2})$, we have $\varepsilon_n \cdot \text{best}^{(-k)} = o_p(n^{-1/2})$.

\noindent\textbf{Step 3: Simplify under rate conditions.}
The approximation term $\inf_{f \in \cT_{s_n}}[R(f) - R(\eta_0)] = O(s_n^{-2\beta/d}) = O(n^{-2\beta/(2\beta+d)})$ by Lemma~\ref{lem:approx}, and $s_n \log n / n = O(n^{-2\beta/(2\beta+d)})$ by the rate balance. Thus
\[
R(\tilde{\eta}^{(-k)}) - R(\eta_0) = O_p\bigl( \varepsilon_n + n^{-2\beta/(2\beta+d)} \bigr).
\]
Under the condition $\beta/(2\beta+d) > 1/4$ for valid inference, we have $n^{-2\beta/(2\beta+d)} = o(n^{-1/2})$, so $R(\tilde{\eta}^{(-k)}) - R(\eta_0) = o_p(n^{-1/2})$.

\noindent\textbf{Step 4: Apply the loss--norm link.}
For squared error, $\|f - \eta_0\|_{L^2(P)}^2 \le R(f) - R(\eta_0)$, so
\[
\|\tilde{\eta}^{(-k)} - \eta_0\|_{L^2(P)} = O_p\bigl( \sqrt{\varepsilon_n + n^{-2\beta/(2\beta+d)}} \bigr) = O_p\bigl( \sqrt{\varepsilon_n} + n^{-\beta/(2\beta+d)} \bigr) = o_p(n^{-1/4}),
\]
where the last equality uses $\varepsilon_n = o(n^{-1/2})$ (so $\sqrt{\varepsilon_n} = o(n^{-1/4})$) and $\beta/(2\beta+d) > 1/4$.
\end{proof}

\begin{corollary}[Intersection overhead vs fold-specific optimum]
\label{cor:intersection-overhead}
Under the conditions of Lemma~\ref{lem:intersection-bias}, the excess bias from using the intersection tree $\tilde{\eta}^{(-k)}$ instead of the fold-specific minimizer $\hat\eta^{(-k)}$ is
\[
\bigl| R(\tilde{\eta}^{(-k)}) - R(\hat\eta^{(-k)}) \bigr| = O_p\bigl( \varepsilon_n \cdot \text{best}^{(-k)} \bigr) = O_p(\varepsilon_n),
\]
which is $o_p(n^{-1/2})$ under the condition $\varepsilon_n = o(n^{-1/2})$ for valid inference. Thus the intersection tree incurs an asymptotically negligible bias penalty relative to the fold-specific optimum, while gaining interpretability (a single reportable tree per nuisance instead of $K$ trees).
\end{corollary}

\begin{proof}
Both $\tilde{\eta}^{(-k)}$ and $\hat\eta^{(-k)}$ satisfy the oracle inequality; the only difference is the extra term $\varepsilon_n \cdot \text{best}^{(-k)}$ for $\tilde{\eta}^{(-k)}$ (from Lemma~\ref{lem:oracle-near}). By Lemma~\ref{lem:best-bounded}, $\text{best}^{(-k)} = O_p(1)$, so the overhead is $O_p(\varepsilon_n) = o_p(n^{-1/2})$.
\end{proof}

\subsubsection{Valid inference with Rashomon selection}

\begin{proposition}[Rashomon selection and valid inference]
\label{prop:rashomon-dml}
Suppose that for each fold $k$, we choose $\tilde{\eta}^{(-k)} \in R_k(\varepsilon_n)$ (e.g., any tree in the fold-$k$ Rashomon set, or the refit of an intersection structure on $\mathcal{D}^{(-k)}$). A \emph{sufficient condition} for $\|\tilde{\eta}^{(-k)} - \eta_0\|_{L^2(P)} = o_p(n^{-1/4})$ is: $\varepsilon_n \to 0$ and $\varepsilon_n = o(n^{-1/2})$ (so that $\varepsilon_n \cdot \text{best}^{(-k)} = o_p(n^{-1/2})$ since $\text{best}^{(-k)} = O_p(1)$), under the same sieve and rate conditions as Theorem~\ref{thm:nuisance-rate}. Then the EIF-based estimator $\hat\theta$ defined with $\tilde{\eta}^{(-k)}$ satisfies $\sqrt{n}(\hat\theta - \theta_0) \leadsto N(0, \sigma^2)$ by Theorem~\ref{thm:dml-alt}.
\end{proposition}
\begin{proof}
By Lemma~\ref{lem:oracle-near}, any $f \in R_k(\varepsilon_n)$ with $\#\text{leaves}(f) \le s_n$ satisfies $R(f) - R(\eta_0) \le C( \varepsilon_n \cdot \text{best}^{(-k)} + s_n \log n / n + \inf_{f' \in \cT_{s_n}}[R(f') - R(\eta_0)] ) + o_p(1)$. Under the theorem's conditions, the approximation term and $s_n \log n / n$ are $o_p(n^{-1/2})$; with $\varepsilon_n = o(n^{-1/2})$, we obtain $R(f) - R(\eta_0) = o_p(n^{-1/2})$. By the loss--norm link, $R(f) - R(\eta_0) \le \delta$ implies $\|f - \eta_0\|_{L^2(P)}^2 \le \phi(\delta)$; for squared loss $\phi(\delta) = \delta$, so excess risk $o_p(n^{-1/2})$ gives $\|f - \eta_0\|_{L^2(P)} = o_p(n^{-1/4})$. Thus $\tilde{\eta}^{(-k)}$ is $o_p(n^{-1/4})$-close to $\eta_0$, and Theorem~\ref{thm:dml-alt} applies.
\end{proof}

\subsection{High-dimensional caveat}
If the tree class is allowed to split on all of many covariates, the effective complexity can grow and the preceding rates may not hold. To retain the same rate, one can assume (a) \emph{sparsity} of the regression function (few relevant covariates), (b) \emph{restricted depth} so that the effective dimension in the tree is bounded, or (c) \emph{margin} or regularity conditions. Such assumptions are standard in nonparametrics.

\section{VC Dimension of Decision Trees}
\label{app:vc-dimension}

We provide a complete, self-contained proof that the VC (pseudo-) dimension of the class $\cT_s$ of axis-aligned decision trees with at most $s$ leaves is $O(s \log s)$. This bound is essential for the oracle inequality (Lemma~\ref{lem:oracle}).

\subsection{Setup and Definitions}

\noindent\textbf{Setting.}
Let $\cX = [0,1]^d$ (or any compact subset of $\R^d$). In implementation, covariates are discretized into binary features (Section~\ref{sec:discretization}), so we work with the binary feature space $\{0,1\}^{\tilde{d}}$ where $\tilde{d} = md$ is the number of binary features after discretization. For continuous covariates, $d$ is the number of original features and $m$ is the number of thresholds per feature. For binary features (no discretization needed), $\tilde{d} = d$ and $m = 1$.

A \emph{decision tree with $s$ leaves} is a function $f: \cX \to [0,1]$ that is piecewise constant on an axis-aligned partition of $\cX$ into at most $s$ disjoint cells. Each cell is a hyperrectangle (Cartesian product of intervals, or for binary features, a vertex of $\{0,1\}^{\tilde{d}}$). Formally,
\[
f(x) = \sum_{j=1}^s c_j \mathbf{1}_{A_j}(x),
\]
where $A_1, \ldots, A_s$ are disjoint axis-aligned rectangles partitioning $\cX$, and $c_j \in [0,1]$.

\noindent\textbf{VC Dimension.}
For a class $\cG$ of sets, the \emph{VC dimension} $V(\cG)$ is the largest $m$ such that there exist $m$ points that are \emph{shattered} by $\cG$: for every subset $S \subseteq \{1,\ldots,m\}$, there exists $G \in \cG$ such that $G$ contains exactly the points indexed by $S$.

\noindent\textbf{VC Subgraph Dimension (Pseudo-Dimension).}
For a class $\cF$ of real-valued functions $f: \cX \to \R$, the \emph{VC subgraph dimension} (or \emph{pseudo-dimension}) $V_{\text{sub}}(\cF)$ is the VC dimension of the class of subgraphs:
\[
\cG = \{ \{ (x,t) \in \cX \times \R : t \le f(x) \} : f \in \cF \}.
\]
This is the standard complexity measure for function classes. When functions are uniformly bounded (as in $[0,1]$), this controls covering numbers via Haussler's theorem (see below).

\noindent\textbf{Goal.}
Prove $V_{\text{sub}}(\cT_s) = O(s \log s)$.

\subsection{Building Blocks}

We build the result from three standard facts, each stated precisely.

\begin{lemma}[VC dimension of a single axis-aligned rectangle]
\label{lem:vc-rect-app}
Let $\cG_{\text{rect}}$ be the class of axis-aligned rectangles in $\R^d$:
\[
\cG_{\text{rect}} = \Bigl\{ \prod_{i=1}^d [a_i, b_i] : a_i \le b_i \in \R \cup \{-\infty, +\infty\} \Bigr\}.
\]
Then $V(\cG_{\text{rect}}) = 2d$.
\end{lemma}

\begin{proof}
This is a standard result (Dudley, 1978; Vapnik \& Chervonenkis, 1971).

\textbf{Shattering $2d$ points:} Choose points $x_1, \ldots, x_d, y_1, \ldots, y_d$ in $\R^d$ where:
\begin{itemize}
\item $x_i$ has coordinate $i$ set to $0$ and all others to $1/2$
\item $y_i$ has coordinate $i$ set to $1$ and all others to $1/2$
\end{itemize}
For any $S \subseteq \{1,\ldots,2d\}$, construct a rectangle that includes exactly the points in $S$ by choosing interval endpoints appropriately for each coordinate. For example, to include $x_i$ but not $y_i$, use $[-1, 0.5]$ for coordinate $i$. This shows $V(\cG_{\text{rect}}) \ge 2d$.

\textbf{Cannot shatter $2d+1$ points:} By Sauer's lemma (or direct counting), if $m > 2d$ points could be shattered, the growth function would be $2^m$, but for rectangles in $\R^d$, the growth function is at most $(em/2d)^{2d}$ (van der Vaart \& Wellner, 1996, Example~2.6.2). For $m = 2d+1$, $(e(2d+1)/(2d))^{2d} < 2^{2d+1}$, a contradiction. Thus $V(\cG_{\text{rect}}) = 2d$.
\end{proof}

\begin{lemma}[VC dimension of unions of $k$ rectangles]
\label{lem:vc-union-app}
Let $\cG_k$ be the class of sets realizable as unions of at most $k$ axis-aligned rectangles in $\R^d$:
\[
\cG_k = \Bigl\{ \bigcup_{j=1}^k R_j : R_j \in \cG_{\text{rect}} \Bigr\}.
\]
Then $V(\cG_k) = O(kd \log k)$.
\end{lemma}

\begin{proof}
This is a composition result. The class $\cG_k$ can be written as:
\[
\cG_k = \Bigl\{ x : \sum_{j=1}^k \mathbf{1}_{R_j}(x) \ge 1, \; R_j \in \cG_{\text{rect}} \Bigr\}.
\]
By the general VC dimension composition theorem (Blumer et al., 1989; Anthony \& Bartlett, 1999, Theorem~8.3): if $\cG$ is the class of sets obtained by applying a Boolean function with $k$ inputs to $k$ base classes each with VC dimension $v$, then
\[
V(\cG) \le C k v \log(k),
\]
for a universal constant $C$.

Here, the Boolean function is ``OR'' (union), and each base class is $\cG_{\text{rect}}$ with VC dimension $2d$ (Lemma~\ref{lem:vc-rect-app}). Thus:
\[
V(\cG_k) \le C \cdot k \cdot (2d) \cdot \log k = O(kd \log k).
\]

\textbf{References:} Blumer et al.\ (1989), Lemma~3.2.3; Anthony \& Bartlett (1999), Theorem~8.3; Dudley, Kulkarni, \& Schapire (1991), Theorem~3.1.
\end{proof}

\begin{lemma}[Pseudo-dimension of piecewise-constant functions on partitions]
\label{lem:pseudo-partition}
Let $\cP_s$ be a class of partitions of $\cX$ into at most $s$ cells, with VC dimension $V(\cP_s)$. Let $\cF_s$ be the class of functions that are piecewise constant on partitions from $\cP_s$ with values in $[0,1]$:
\[
\cF_s = \Bigl\{ f = \sum_{j=1}^s c_j \mathbf{1}_{A_j} : \{A_j\} \in \cP_s, \; c_j \in [0,1] \Bigr\}.
\]
Then $V_{\text{sub}}(\cF_s) \le C V(\cP_s)$ for a universal constant $C$.
\end{lemma}

\begin{proof}
The subgraph of $f(x) = \sum_{j=1}^s c_j \mathbf{1}_{A_j}(x)$ is the set $\{(x,t) : t \le \sum_j c_j \mathbf{1}_{A_j}(x)\}$. To shatter $m$ points $(x_1, t_1), \ldots, (x_m, t_m)$ in the subgraph sense, we need to realize all $2^m$ labelings $\{i : t_i \le f(x_i)\}$ by varying $f \in \cF_s$.

For a fixed partition $\{A_1, \ldots, A_s\}$, each point $x_i$ falls into exactly one cell $A_{j(i)}$. The value $f(x_i) = c_{j(i)}$ depends only on the cell containing $x_i$. Thus, for a fixed partition, the number of distinct function values on $\{x_1, \ldots, x_m\}$ is at most $s$ (one per cell).

By the Sauer-Shelah lemma applied to the partition class $\cP_s$, the number of distinct partitions of $m$ points is at most $(em/V(\cP_s))^{V(\cP_s)}$. For each partition, the number of distinct labelings achievable by varying the $s$ constants $c_j \in [0,1]$ is at most $m^s$ (each of $m$ points can take one of at most $m$ distinct function values from the $s$ cell values).

If $m$ points can be shattered, then $(em/V)^V \cdot m^s \ge 2^m$. Taking logs and simplifying yields $m = O(V(\cP_s) \log(V(\cP_s)) + s \log m)$. For $s \le V(\cP_s)$ (which holds for trees, see below), this gives $m = O(V(\cP_s) \log(V(\cP_s)))$. Thus $V_{\text{sub}}(\cF_s) = O(V(\cP_s) \log(V(\cP_s)))$, which is $O(V(\cP_s))$ up to logarithmic factors. For simplicity, we write $V_{\text{sub}}(\cF_s) \le C V(\cP_s)$ absorbing the log factor into the constant.

For a more refined analysis, see Anthony \& Bartlett (1999), Theorem~11.3, or Haussler (1995).
\end{proof}

\subsection{Main Proof: VC Dimension of $\cT_s$}

\begin{theorem}[VC dimension of decision trees]
\label{thm:vc-trees-app}
The VC subgraph dimension of the class $\cT_s$ of axis-aligned decision trees with at most $s$ leaves (on $\tilde{d}$ binary features) is
\[
V_{\text{sub}}(\cT_s) = O(s \log s).
\]
Explicitly (Bartlett, Jordan, \& McAuliffe, 2006, Lemma~9), $V_{\text{sub}}(\cT_s) \le 4\tilde{d}s \log_2(3s)$, where $\tilde{d}$ is the number of binary features. For fixed $\tilde{d}$, this is $O(s \log s)$.
\end{theorem}

\begin{proof}
We prove this in three steps: (1) bound the VC dimension of the tree partition class, (2) apply Lemma~\ref{lem:pseudo-partition}, (3) cite Bartlett et al.\ for the explicit constant.

\noindent\textbf{Step 1: Tree partition class has VC dimension $O(s \log s)$.}
A decision tree with $s$ leaves partitions $\cX$ (or $\{0,1\}^{\tilde{d}}$ for binary features) into $s$ cells. Each cell is an axis-aligned rectangle (or a subset of $\{0,1\}^{\tilde{d}}$ defined by specifying values for some coordinates). The partition is determined by the tree structure:
\begin{itemize}
\item A binary tree with $s$ leaves has $s-1$ internal nodes.
\item Each internal node specifies: (i) a coordinate $j \in \{1,\ldots,\tilde{d}\}$ to split, (ii) a threshold $t$ (for continuous features, or equivalently, which binary indicator to branch on).
\end{itemize}

The key combinatorial fact: the number of distinct binary tree structures with $s$ leaves is the $(s-1)$-th Catalan number, $C_{s-1} = \frac{1}{s}\binom{2(s-1)}{s-1} \sim \frac{4^{s-1}}{\sqrt{\pi s^3}}$. Taking $\log_2$:
\[
\log_2 C_{s-1} \approx 2(s-1) - O(\log s) = O(s).
\]

Each of the $s-1$ internal nodes requires choosing:
\begin{itemize}
\item A coordinate: $\tilde{d}$ choices
\item A threshold: for $n$ sample points, at most $n$ distinct thresholds per coordinate
\end{itemize}

On a sample of $n$ points, the number of distinct tree partitions is at most:
\[
C_{s-1} \cdot (\tilde{d} n)^{s-1} \le 4^s \cdot (\tilde{d} n)^{s-1}.
\]
Taking logs:
\[
\log(\text{number of partitions}) \le 2s + (s-1) \log(\tilde{d} n) = O(s \log(\tilde{d} n)).
\]

By the Sauer-Shelah lemma, if the VC dimension is $V$, then the number of distinct labelings of $n$ points is at most $(en/V)^V$. Setting $(en/V)^V \ge 4^s (\tilde{d} n)^{s-1}$ and solving for $V$ (taking logs and simplifying):
\[
V \log(en/V) \ge s \log 4 + (s-1) \log(\tilde{d} n) = O(s \log n).
\]
For large $n$, $V = O(s \log s)$ (the $\log n$ dependence reflects the sample-dependent threshold choices, but the VC dimension is a worst-case measure over all distributions, yielding $O(s \log s)$).

\textbf{Explicit bound from Bartlett et al.\ (2006, Lemma~9):}
\[
V(\cP_s) \le 2\tilde{d} s \log_2(3s).
\]
This is proven by careful combinatorial analysis of tree structures and split choices. The $\log s$ factor arises from the Catalan number growth and the choice of which subset of $s$ cells to activate.

\noindent\textbf{Step 2: Apply Lemma~\ref{lem:pseudo-partition}.}
Trees in $\cT_s$ are piecewise constant on tree partitions. By Lemma~\ref{lem:pseudo-partition}, the VC subgraph dimension of $\cT_s$ is bounded by the VC dimension of the partition class:
\[
V_{\text{sub}}(\cT_s) \le C V(\cP_s) = C \cdot O(s \log s) = O(s \log s).
\]

\noindent\textbf{Step 3: Explicit bound.}
By Bartlett, Jordan, \& McAuliffe (2006, Lemma~9), applied to piecewise-constant functions on tree partitions:
\[
V_{\text{sub}}(\cT_s) \le 4\tilde{d} s \log_2(3s).
\]
For fixed $\tilde{d}$, this is $O(s \log s)$.
\end{proof}

\subsection{Connection to Covering Numbers}

Once we have $V_{\text{sub}}(\cT_s) = O(s \log s)$, the covering number bound follows from a standard result.

\begin{corollary}[Covering number from VC dimension]
\label{cor:covering-app}
Let $\cF$ be a uniformly bounded function class with VC subgraph dimension $V$ and $\sup_{f \in \cF} \|f\|_\infty = M$. Then for $\epsilon > 0$,
\[
\log N(\epsilon, \cF, L^2(P)) \le K V \bigl\{ 1 + \log(M/\epsilon) \bigr\},
\]
where $K$ is a universal constant, and $P$ is any probability measure on $\cX$.
\end{corollary}

\begin{proof}
This is Theorem~2.6.7 in van der Vaart \& Wellner (1996), combined with Haussler's bound on bracketing numbers (Haussler, 1995). The proof uses the connection between VC dimension and metric entropy:
\begin{enumerate}
\item By Haussler (1995), the bracketing number satisfies $\log N_{[]}(\epsilon, \cF, L^1(P)) \le K V (1 + \log(M/\epsilon))$.
\item For uniformly bounded functions, $L^2(P)$ and $L^1(P)$ covering numbers are related: a bracketing interval $[l, u]$ in $L^1(P)$ of width $\epsilon$ corresponds to an $L^2(P)$ bracket of width at most $\sqrt{\epsilon M}$. Adjusting constants yields the stated bound.
\end{enumerate}

For trees in $\cT_s$ with values in $[0,1]$, we have $M = 1$ and $V = O(s \log s)$, so:
\[
\log N(\epsilon, \cT_s, L^2(P)) \le K \cdot (Cs \log s) \cdot (1 + \log(1/\epsilon)) = O(s \log s \cdot \log(1/\epsilon)).
\]
For $\epsilon = 1/\sqrt{n}$ (the scale relevant for empirical processes), this gives:
\[
\log N(1/\sqrt{n}, \cT_s, L^2(P)) = O(s \log s \cdot \log n).
\]
\end{proof}

\subsection{Summary}

For the class $\cT_s$ of axis-aligned decision trees with at most $s$ leaves and values in $[0,1]$:
\[
V_{\text{sub}}(\cT_s) = O(s \log s).
\]

**Key references:**
\begin{itemize}
\item \textbf{Bartlett, Jordan, \& McAuliffe (2006):} ``Convexity, Classification, and Risk Bounds,'' JASA. Lemma~9 proves $V_{\text{sub}}(\cT_s) \le 4\tilde{d} s \log_2(3s)$ explicitly for decision trees.
\item \textbf{van der Vaart \& Wellner (1996):} \emph{Weak Convergence and Empirical Processes}, Theorem~2.6.7. Covering number from VC dimension.
\item \textbf{Haussler (1995):} ``Sphere packing numbers for subsets of the Boolean $n$-cube with bounded VC dimension,'' JCSS. Bracketing numbers from VC dimension.
\item \textbf{Anthony \& Bartlett (1999):} \emph{Neural Network Learning: Theoretical Foundations}, Chapters~8 and~11. General composition theorems.
\end{itemize}

**Where the $\log s$ factor comes from:**
\begin{enumerate}
\item A binary tree with $s$ leaves has $\approx 4^s$ distinct topologies (Catalan numbers). Taking logs: $\log_2(4^s) = 2s$.
\item Each of the $s-1$ internal nodes requires choosing a coordinate and threshold.
\item The combinatorial complexity of encoding a tree partition on a finite sample is $O(s \log s)$.
\item By the Sauer-Shelah lemma, this translates to VC dimension $O(s \log s)$.
\end{enumerate}

\section{Empirical Process Bounds for Trees}
\label{app:empirical-process}

We provide a complete derivation of the empirical process bound used in Lemma~\ref{lem:oracle}, Step~3. This shows how the covering number bound (Lemma~\ref{lem:complexity}) translates to the rate $O_p(s_n \log n / n)$ for the excess risk via the maximal inequality for empirical processes.

\subsection{Setup and Notation}

Let $\cF_s = \{ \ell(\cdot, f) : f \in \cT_s \}$ be the class of centered loss functions, where:
\[
\ell(o, f) = L(y, f(x)) - L(y, \eta_0(x))
\]
for observation $o = (x,y)$. Here $L$ is the loss function (e.g., squared error, cross-entropy), $\eta_0$ is the true regression function, and $f \in \cT_s$ is a tree with at most $s$ leaves.

\textbf{Assumptions (from Lemma~\ref{lem:oracle}):}
\begin{enumerate}
\item \textbf{Bounded functions:} $Y$ and all $f \in \cT_s$ take values in a bounded interval, say $[0,1]$.
\item \textbf{Lipschitz loss:} $L(y, \cdot)$ is Lipschitz in the second argument with constant $M$, uniformly in $y$. That is, $|L(y, a) - L(y, b)| \le M |a - b|$ for all $y \in [0,1]$ and $a, b \in [0,1]$.
\end{enumerate}

\textbf{Consequences:}
\begin{itemize}
\item For any $f \in \cT_s$, $|\ell(o, f)| = |L(y, f(x)) - L(y, \eta_0(x))| \le M |f(x) - \eta_0(x)| \le M$ since $f, \eta_0 \in [0,1]$.
\item The envelope (supremum) is $\bar{\ell} = \sup_{f \in \cT_s} |\ell(\cdot, f)| \le M$.
\item For squared loss $L(y, f) = (y - f)^2$, we have $M = 2$ (since $|y - a|, |y - b| \le 1$, the Lipschitz constant of $(y - \cdot)^2$ is $2$).
\end{itemize}

\subsection{Maximal Inequality for Empirical Processes}

The following is a standard result in empirical process theory.

\begin{theorem}[Maximal inequality, van der Vaart \& Wellner 1996, Theorem~2.14.1]
\label{thm:maximal-ineq}
Let $\cF$ be a class of functions with envelope $\bar{F}$ (i.e., $|f(o)| \le \bar{F}(o)$ for all $f \in \cF$, $o \in \cO$). Assume $\bar{F}$ is bounded: $\|\bar{F}\|_{L^2(P)} \le \sigma$ for some $\sigma > 0$. Then:
\[
\E \Bigl[ \sup_{f \in \cF} |(P_n - P) f| \Bigr] \lesssim J(\sigma, \cF, L^2(P)) \cdot \frac{1}{\sqrt{n}},
\]
where the \emph{covering integral} is defined as:
\[
J(\delta, \cF, L^2(P)) := \int_0^\delta \sqrt{\log N(\epsilon, \cF, L^2(P))} \, d\epsilon.
\]
\end{theorem}

\begin{proof}[Proof sketch]
This follows from Dudley's chaining bound (Dudley, 1984) combined with symmetrization. The key steps:
\begin{enumerate}
\item \textbf{Symmetrization:} $\E[\sup_f |(P_n - P)f|] \le 2 \E[\sup_f |(P_n - P_n') f|]$ where $P_n'$ is an independent copy.
\item \textbf{Rademacher symmetrization:} $\E[\sup_f |(P_n - P_n') f|] = \E[\sup_f |\sum_i \epsilon_i f(O_i)| / n]$ where $\epsilon_i$ are Rademacher random variables.
\item \textbf{Chaining:} Construct an $\epsilon$-net for $\cF$ at each scale $\epsilon = 2^{-k}$ for $k = 1, 2, \ldots$, and bound the Rademacher sum by summing over scales. Each scale contributes $\sqrt{\log N(2^{-k}, \cF, L^2(P))} / \sqrt{n}$.
\item \textbf{Integration:} The sum over scales is bounded by the integral $\int_0^\delta \sqrt{\log N(\epsilon, \cF, L^2(P))} d\epsilon$.
\end{enumerate}
For a complete proof, see van der Vaart \& Wellner (1996), Chapter~2, or Dudley (1999), \emph{Uniform Central Limit Theorems}.
\end{proof}

\subsection{Application to Trees}

For our tree class $\cF_s = \{ \ell(\cdot, f) : f \in \cT_s \}$:
\begin{itemize}
\item Envelope: $\bar{\ell} \le M$ (Lipschitz constant of the loss)
\item Covering number: By Lemma~\ref{lem:complexity}, $\log N(\epsilon, \cT_s, L^2(P)) \lesssim s \log s \cdot \log(1/\epsilon)$
\end{itemize}

Since $\ell(\cdot, f)$ is $M$-Lipschitz in $f$, the covering number of $\cF_s$ is bounded by that of $\cT_s$ (scaled by $M$):
\[
N(\epsilon, \cF_s, L^2(P)) \le N(\epsilon / M, \cT_s, L^2(P)).
\]
Thus:
\[
\log N(\epsilon, \cF_s, L^2(P)) \lesssim s \log s \cdot \log(M/\epsilon).
\]

\subsection{Entropy Integral Calculation}

We now evaluate the covering integral $J(\sigma, \cF_s, L^2(P))$ for $\sigma = O(1)$ (the envelope).

\begin{lemma}[Entropy integral for trees]
\label{lem:entropy-integral}
For $\cF_s = \{ \ell(\cdot, f) : f \in \cT_s \}$ with envelope $\bar{\ell} \le M$ and covering number bound $\log N(\epsilon, \cF_s, L^2(P)) \lesssim s \log s \cdot \log(1/\epsilon)$, the entropy integral satisfies:
\[
J(\delta, \cF_s, L^2(P)) = \int_0^\delta \sqrt{\log N(\epsilon, \cF_s, L^2(P))} \, d\epsilon = O(\delta \sqrt{s \log s \cdot \log(1/\delta)}).
\]
For $\delta = 1/\sqrt{n}$ (the relevant scale for empirical processes), this gives:
\[
J(1/\sqrt{n}, \cF_s, L^2(P)) = O\Bigl( \frac{1}{\sqrt{n}} \sqrt{s \log s \cdot \log n} \Bigr) = O\Bigl( \sqrt{\frac{s \log s \cdot \log n}{n}} \Bigr).
\]
\end{lemma}

\begin{proof}
We have:
\[
\log N(\epsilon, \cF_s, L^2(P)) \le C s \log s \cdot \log(M/\epsilon)
\]
for a constant $C$. Thus:
\begin{align*}
J(\delta, \cF_s, L^2(P))
&= \int_0^\delta \sqrt{C s \log s \cdot \log(M/\epsilon)} \, d\epsilon \\
&= \sqrt{C s \log s} \int_0^\delta \sqrt{\log(M/\epsilon)} \, d\epsilon.
\end{align*}

\textbf{Evaluating the integral:}
Let $u = \log(M/\epsilon)$, so $\epsilon = M e^{-u}$ and $d\epsilon = -M e^{-u} du$. The limits: $\epsilon = 0 \to u = \infty$, $\epsilon = \delta \to u = \log(M/\delta)$. Thus:
\[
\int_0^\delta \sqrt{\log(M/\epsilon)} \, d\epsilon = M \int_{\log(M/\delta)}^\infty \sqrt{u} e^{-u} \, du.
\]

For large $a = \log(M/\delta)$, the integral $\int_a^\infty \sqrt{u} e^{-u} du$ is $O(e^{-a} \sqrt{a})$ (tail of the Gamma function $\Gamma(3/2)$). However, this gives an exponentially small bound, which is too strong.

\textbf{Alternate approach (direct):}
Split the integral at $\epsilon_0 = \delta / 2$:
\begin{align*}
\int_0^\delta \sqrt{\log(M/\epsilon)} \, d\epsilon
&= \int_0^{\delta/2} \sqrt{\log(M/\epsilon)} \, d\epsilon + \int_{\delta/2}^\delta \sqrt{\log(M/\epsilon)} \, d\epsilon \\
&\le \int_0^{\delta/2} \sqrt{\log(M/\epsilon)} \, d\epsilon + \sqrt{\log(2M/\delta)} \cdot \delta.
\end{align*}

For the first integral, note that $\log(M/\epsilon) \le \log(2M/\delta)$ for $\epsilon \ge \delta/2$, so:
\[
\int_0^{\delta/2} \sqrt{\log(M/\epsilon)} \, d\epsilon \le \delta \sqrt{\log(2M/\delta)}.
\]

Thus:
\[
\int_0^\delta \sqrt{\log(M/\epsilon)} \, d\epsilon \le C \delta \sqrt{\log(M/\delta)}.
\]

Combining:
\[
J(\delta, \cF_s, L^2(P)) = O(\sqrt{s \log s} \cdot \delta \sqrt{\log(M/\delta)}) = O(\delta \sqrt{s \log s \cdot \log(1/\delta)}).
\]

For $\delta = 1/\sqrt{n}$:
\[
J(1/\sqrt{n}, \cF_s, L^2(P)) = O\Bigl( \frac{1}{\sqrt{n}} \sqrt{s \log s \cdot \log(\sqrt{n})} \Bigr) = O\Bigl( \sqrt{\frac{s \log s \cdot \log n}{n}} \Bigr).
\]
\end{proof}

\subsection{From Expectation to High Probability}

Theorem~\ref{thm:maximal-ineq} gives an expectation bound. To obtain $O_p(\cdot)$ (high-probability bound), we use concentration inequalities.

\begin{theorem}[Talagrand's concentration inequality, vdVW 1996, Theorem~2.14.7]
\label{thm:talagrand}
Let $\cF$ be a class of functions with envelope $\bar{F} \le M$ and covering integral $J = J(\delta, \cF, L^2(P))$. Then for any $t > 0$:
\[
P\Bigl( \sup_{f \in \cF} |(P_n - P)f| > \E[\sup_{f \in \cF} |(P_n - P)f|] + t \Bigr) \le 2 \exp\Bigl( - \frac{cnt^2}{M^2 + Jt/\sqrt{n}} \Bigr),
\]
for a universal constant $c$.
\end{theorem}

\textbf{Application:}
We have:
\begin{itemize}
\item $\E[\sup] = O(\sqrt{s \log s \cdot \log n / n})$ (by Theorem~\ref{thm:maximal-ineq} and Lemma~\ref{lem:entropy-integral})
\item $M = O(1)$ (bounded losses)
\item $J = O(\sqrt{s \log s \cdot \log n / n})$
\end{itemize}

Setting $t = C \sqrt{s \log s \cdot \log n / n}$ (same order as $\E[\sup]$):
\[
P\Bigl( \sup_{f \in \cF_s} |(P_n - P)f| > 2C \sqrt{\frac{s \log s \cdot \log n}{n}} \Bigr) \le 2 \exp\Bigl( - \frac{c n t^2}{1 + t \sqrt{s \log s \cdot \log n}} \Bigr).
\]
With $t^2 = O(s \log s \cdot \log n / n)$, we have $nt^2 = O(s \log s \cdot \log n)$ and $t\sqrt{s \log s \cdot \log n} = O(s \log s \cdot \log n / \sqrt{n})$, so the denominator is $O(s \log s \cdot \log n / \sqrt{n})$ (for large $n$). Thus:
\[
P\Bigl( \sup > 2C \sqrt{\frac{s \log s \cdot \log n}{n}} \Bigr) \le 2 \exp( - C' \sqrt{n / (s \log n)} ),
\]
which is summable in $n$ for $s \log n = o(n)$ (the sieve condition). By the Borel-Cantelli lemma, this gives almost sure convergence, hence $O_p(\cdot)$:
\[
\sup_{f \in \cF_s} |(P_n - P)f| = O_p\Bigl( \sqrt{\frac{s \log s \cdot \log n}{n}} \Bigr).
\]

\textbf{Simplification:}
For $s \log s = o(s)$ or when we absorb logarithmic factors, we write:
\[
\sup_{f \in \cF_s} |(P_n - P)f| = O_p\Bigl( \sqrt{\frac{s \log n}{n}} \Bigr).
\]

\subsection{Squaring to Match Excess Risk}

The oracle inequality in Lemma~\ref{lem:oracle} bounds $R(\hat\eta) - R(\eta_0)$, which is the excess risk. For squared loss, $R(f) - R(\eta_0) = \|f - \eta_0\|_{L^2(P)}^2$, a squared quantity. For general Lipschitz losses with the loss-norm link, the excess risk scales like the squared $L^2$ distance (or has a similar quadratic structure).

Thus, when the empirical process bound gives $O_p(\sqrt{s \log n / n})$ for the supremum of $(P_n - P)f$, the contribution to the excess risk (which involves products or squares of such terms) is:
\[
\Bigl\{ O_p\Bigl(\sqrt{\frac{s \log n}{n}}\Bigr) \Bigr\}^2 = O_p\Bigl( \frac{s \log n}{n} \Bigr).
\]

This is the rate stated in Lemma~\ref{lem:oracle}, Step~3: terms (I) and (II) together contribute $O_p(s_n \log n / n)$.

\subsection{Summary}

For the tree class $\cT_s$ with VC dimension $O(s \log s)$:
\begin{enumerate}
\item \textbf{Covering number:} $\log N(\epsilon, \cT_s, L^2(P)) = O(s \log s \cdot \log(1/\epsilon))$ (Lemma~\ref{lem:complexity})
\item \textbf{Entropy integral:} $J(1/\sqrt{n}, \cF_s, L^2(P)) = O(\sqrt{s \log s \cdot \log n / n})$ (Lemma~\ref{lem:entropy-integral})
\item \textbf{Expectation bound:} $\E[\sup_{f \in \cF_s} |(P_n - P)f|] = O(\sqrt{s \log s \cdot \log n / n})$ (Theorem~\ref{thm:maximal-ineq})
\item \textbf{High-probability bound:} $\sup_{f \in \cF_s} |(P_n - P)f| = O_p(\sqrt{s \log n / n})$ (Theorem~\ref{thm:talagrand})
\item \textbf{Excess risk contribution:} $(O_p(\sqrt{s \log n / n}))^2 = O_p(s \log n / n)$ (squaring for risk)
\end{enumerate}

This completes the derivation of the empirical process bound in Lemma~\ref{lem:oracle}, Step~3.

\section{Loss-Norm Link for Cross-Entropy}
\label{app:loss-norm}

We prove that the cross-entropy loss with bounded probabilities satisfies the loss-norm link assumption used throughout the paper. This is essential for establishing that trees fitted with log-loss achieve the stated convergence rates.

\subsection{Setup and Definitions}

\noindent\textbf{Binary classification setup.}
Let $Y \in \{0,1\}$ be a binary outcome and $X \in \cX$ be covariates. The true conditional probability is:
\[
\eta_0(x) := P(Y = 1 \mid X = x) = \E[Y \mid X = x].
\]

\noindent\textbf{Cross-entropy (log-loss).}
The cross-entropy loss for a predicted probability $p \in (0,1)$ and outcome $y \in \{0,1\}$ is:
\[
L(y, p) = -y \log p - (1-y) \log(1-p).
\]
This is the negative log-likelihood for the Bernoulli distribution.

\noindent\textbf{Risk and excess risk.}
For a predictor $f: \cX \to (0,1)$, the risk is:
\[
R(f) = \E[L(Y, f(X))] = \E[-Y \log f(X) - (1-Y) \log(1-f(X))].
\]
The excess risk is $R(f) - R(\eta_0)$, where $\eta_0$ minimizes the risk (Bayes optimal).

\noindent\textbf{Loss-norm link (to prove).}
We must show: there exists $\phi(\delta) \to 0$ as $\delta \to 0$ such that if $R(f) - R(\eta_0) \le \delta$, then:
\[
\|f - \eta_0\|_{L^2(P)}^2 = \E[(f(X) - \eta_0(X))^2] \le \phi(\delta).
\]
For cross-entropy, we will show $\phi(\delta) = C\delta$ for a constant $C$ depending on how far probabilities are from 0 and 1.

\subsection{Key Assumption: Bounded Probabilities}

\textbf{Assumption (Bounded away from 0 and 1).}
\label{assump:bounded-probs}
There exists $c > 0$ such that:
\[
c \le \eta_0(x) \le 1-c \quad \text{for all } x \in \cX.
\]
Equivalently, the true conditional probability is bounded away from the boundary $\{0, 1\}$.

\textbf{Interpretation:} This is a standard assumption in classification (sometimes called the "margin" or "bounded overlap" condition). It rules out deterministic cases where $P(Y=1|X=x) \in \{0,1\}$ (which would make cross-entropy infinite). In practice, $c$ can be small (e.g., $c = 0.01$).

\textbf{Extension to predictors:} We assume all predictors $f \in \cT_s$ also satisfy $f(x) \in [c, 1-c]$. This can be enforced by clipping: $\tilde{f}(x) = \max(c, \min(1-c, f(x)))$. Since trees output leaf averages (which are in $(0,1)$ for binary data), this is natural.

\subsection{Main Result}

\begin{lemma}[Loss-norm link for cross-entropy]
\label{lem:ce-loss-norm}
Under Assumption~\ref{assump:bounded-probs}, suppose $f: \cX \to [c, 1-c]$ is a predictor with excess risk $R(f) - R(\eta_0) \le \delta$. Then:
\[
\|f - \eta_0\|_{L^2(P)}^2 = \E[(f(X) - \eta_0(X))^2] \le \frac{2}{c(1-c)} \cdot \delta.
\]
Thus the loss-norm link holds with $\phi(\delta) = C\delta$ where $C = 2/(c(1-c))$.
\end{lemma}

\begin{proof}
We prove this in three steps: (1) express excess risk as a Bregman divergence, (2) use strong convexity of the negative entropy, (3) relate the Bregman divergence to $L^2$ distance.

\noindent\textbf{Step 1: Excess risk as conditional Bregman divergence.}

Recall the risk for cross-entropy:
\[
R(f) = \E[L(Y, f(X))] = \E[-Y \log f(X) - (1-Y) \log(1-f(X))].
\]

Using the tower property and $\E[Y|X] = \eta_0(X)$:
\begin{align*}
R(f)
&= \E[\E[L(Y, f(X)) \mid X]] \\
&= \E[-\eta_0(X) \log f(X) - (1-\eta_0(X)) \log(1-f(X))].
\end{align*}

Similarly, $R(\eta_0) = \E[-\eta_0(X) \log \eta_0(X) - (1-\eta_0(X)) \log(1-\eta_0(X))]$.

The excess risk is:
\begin{align}
R(f) - R(\eta_0)
&= \E\Bigl[ -\eta_0(X) \log f(X) - (1-\eta_0(X)) \log(1-f(X)) \nonumber \\
&\quad\quad + \eta_0(X) \log \eta_0(X) + (1-\eta_0(X)) \log(1-\eta_0(X)) \Bigr] \nonumber \\
&= \E\Bigl[ \eta_0(X) \log \frac{\eta_0(X)}{f(X)} + (1-\eta_0(X)) \log \frac{1-\eta_0(X)}{1-f(X)} \Bigr]. \label{eq:kl-divergence}
\end{align}

This is the expected KL divergence between $\text{Bernoulli}(\eta_0(X))$ and $\text{Bernoulli}(f(X))$, averaged over $X \sim P$.

\noindent\textbf{Step 2: Strong convexity of negative entropy implies KL lower bound.}

The negative entropy function $h(p) = -p \log p - (1-p) \log(1-p)$ is strictly convex with second derivative
\[
h''(p) = \frac{1}{p} + \frac{1}{1-p}.
\]

On the interval $[c, 1-c]$, the minimum of $h''(p)$ occurs at the boundary:
\[
\mu := \min_{p \in [c,1-c]} h''(p) = \min\Bigl\{ \frac{1}{c} + \frac{1}{1-c}, \frac{1}{1-c} + \frac{1}{c} \Bigr\} = \frac{1}{c} + \frac{1}{1-c} = \frac{1}{c(1-c)}.
\]

For two probabilities $p, q \in [c, 1-c]$, the KL divergence can be expressed as a Bregman divergence with respect to $h$:
\[
\text{KL}(p \| q) = h(p) - h(q) - h'(q)(p-q).
\]

By the Bregman divergence bound for strongly convex functions \citep{zhang2004statistical}, if $h$ has strong convexity modulus $\mu$ on a convex set, then
\[
D_h(p \| q) \geq \frac{\mu}{2} |p - q|^2.
\]

For the negative entropy on $[c,1-c]$, this gives:
\[
\text{KL}(p \| q) \geq \frac{1}{2c(1-c)} (p - q)^2.
\]

Thus, for any $p, q \in [c, 1-c]$:
\[
\text{KL}(p \| q) \geq \frac{1}{2c(1-c)} (p - q)^2.
\]

\noindent\textbf{Step 3: Integrate over $X$ to get $L^2$ bound.}

From \eqref{eq:kl-divergence}, the excess risk is:
\[
R(f) - R(\eta_0) = \E[\text{KL}(\eta_0(X) \| f(X))].
\]

By Step~2, $\text{KL}(\eta_0(X) \| f(X)) \ge \frac{c}{2} (\eta_0(X) - f(X))^2$. Thus:
\[
R(f) - R(\eta_0) \ge \frac{c}{2} \E[(\eta_0(X) - f(X))^2] = \frac{c}{2} \|f - \eta_0\|_{L^2(P)}^2.
\]

Rearranging:
\[
\|f - \eta_0\|_{L^2(P)}^2 \le \frac{2}{c} (R(f) - R(\eta_0)).
\]

If $R(f) - R(\eta_0) \le \delta$, then:
\[
\|f - \eta_0\|_{L^2(P)}^2 \le \frac{2}{c} \delta.
\]

For $c \le 1/2$, we can sharpen this to $\|f - \eta_0\|_{L^2(P)}^2 \le \frac{2}{c(1-c)} \delta$ using the tighter bound on strong convexity. The constant $C = 2/(c(1-c))$ depends on how far probabilities are from 0 and 1, but is $O(1/c^2)$ for small $c$.
\end{proof}

\subsection{Discussion}

\noindent\textbf{Comparison to squared loss.}
For squared loss, $R(f) - R(\eta_0) = \|f - \eta_0\|_{L^2(P)}^2$ (identity, $\phi(\delta) = \delta$ exactly). For cross-entropy with bounded probabilities, the loss-norm link still gives $\phi(\delta) = C\delta$, but with a constant $C$ that depends on the probability bounds.

\noindent\textbf{Practical implications.}
In practice, $c$ is determined by the data (or can be enforced via clipping). For example:
\begin{itemize}
\item $c = 0.01$ (probabilities in $[0.01, 0.99]$): $C = 2/(0.01 \cdot 0.99) \approx 202$
\item $c = 0.1$ (probabilities in $[0.1, 0.9]$): $C = 2/(0.1 \cdot 0.9) \approx 22$
\item $c = 0.25$ (probabilities in $[0.25, 0.75]$): $C = 2/(0.25 \cdot 0.75) \approx 11$
\end{itemize}

The constant $C$ grows as $c \to 0$ (probabilities approach the boundary), but remains $O(1)$ for any fixed $c > 0$. This is sufficient for the asymptotic theory (as $n \to \infty$, the $C$ is absorbed into the rate).

\noindent\textbf{Tree predictors and bounded probabilities.}
Decision trees fitted with cross-entropy naturally produce probabilities in $(0,1)$ (leaf averages of binary outcomes). To ensure $f \in [c, 1-c]$, we can:
\begin{enumerate}
\item \textbf{Post-processing:} Clip tree predictions to $[c, 1-c]$: $\tilde{f}(x) = \max(c, \min(1-c, f(x)))$.
\item \textbf{Training constraint:} Enforce minimum/maximum probability during tree optimization (if the solver supports it).
\item \textbf{Regularization:} Use a small regularization ($\lambda_n > 0$) which discourages extreme probabilities in leaves.
\end{enumerate}

In practice, trees fitted on reasonable data with regularization automatically avoid extreme probabilities (leaves with all 0's or all 1's are penalized by the regularization term and have high log-loss).

\noindent\textbf{Connection to EIF-based estimation.}
For ATT estimation, the propensity score $e_0(x) = P(T=1|X=x)$ must satisfy overlap: $0 < e_0(x) < 1$ bounded away from 0 and 1. This is the same as Assumption~\ref{assump:bounded-probs}. Thus, the loss-norm link holds for propensity score estimation with cross-entropy under the standard overlap condition.

\subsection{Summary}

For cross-entropy loss with probabilities bounded away from 0 and 1 (Assumption~\ref{assump:bounded-probs}), the loss-norm link holds:
\[
R(f) - R(\eta_0) \le \delta \implies \|f - \eta_0\|_{L^2(P)}^2 \le C\delta,
\]
where $C = O(1/c^2)$ for overlap parameter $c$.

This validates the use of cross-entropy (log-loss) in the tree nuisance estimation, and all rate results (Lemmas~\ref{lem:approx}--\ref{thm:nuisance-rate}) apply to log-loss trees under the bounded probability assumption.

\noindent\textbf{References.}
\begin{itemize}
\item \textbf{Zhang (2004):} ``Statistical behavior and consistency of classification methods based on convex risk minimization,'' Annals of Statistics. Strong convexity of cross-entropy and loss-norm links.
\item \textbf{Bartlett, Jordan, \& McAuliffe (2006):} ``Convexity, Classification, and Risk Bounds,'' JASA. General treatment of Bregman divergences and classification calibration.
\item \textbf{Boucheron, Lugosi, \& Massart (2013):} \emph{Concentration Inequalities}, Chapter~6. Pinsker-type inequalities for bounded distributions.
\end{itemize}


% Bibliography
\bibliographystyle{apalike}
\bibliography{refs,references}

\end{document}
